--- 
title_meta  : Chapter 1
title       : Intro to basics
description : In this chapter, you will take your first steps with R. You will learn how to use the console as a calculator and how to assign variables. You will also get to know the basic data types in R. Let us get started!
framework   : datacamp
mode        : selfcontained
---


## How it works

In the editor on the right you should type R code to solve the exercises. When you hit the 'Submit Answer' button, every line of code is interpreted and executed by R and you get a message whether or not your code was correct. The output of your R code is shown in the console in the lower right corner.

R makes use of the `#` sign to add comments, so that you and others can understand what the R code is about. Just like Twitter! Comments are not run as R code, so they will not influence your result. For example, _Calculate 3 + 4_ in the editor on the right is a comment.

You can also execute R commands straight in the console. This is a good way to experiment with R code, as your submission is not checked for correctness.

*** =instructions
- In the editor on the right there is already some sample code. Can you see which lines are actual R code and which are comments?
- Add a line of code that calculates the sum of 6 and 12, and hit the 'Submit Answer' button.

*** =hint
Just click the 'Submit Answer' button.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Calculate 3 + 4
3 + 4

# Calculate 6 + 12

```

*** =solution
```{r eval=FALSE}
# Calculate 3 + 4
3 + 4

# Calculate 6 + 12
6 + 12
```

*** =sct
```{r eval=FALSE}
test_output_contains("7", incorrect_msg = "Do not remove the line of R code that calculates the sum of 3 and 4. Instead, just add another line that calculates the sum of 5 and 12.")
test_output_contains("18", incorrect_msg = "Make sure to add a line of R code, that calculates the sum of 5 and 12. Do not start the line with a <code>#</code>, otherwise your R code is not executed!")
success_msg("Awesome! See how the console shows the result of the R code you submitted? Now that you're familiar with the interface, let's get down to R business!")
```

---

## Little arithmetics with R

As previously stated, in its most basic form R can be used as a simple calculator. Consider the following arithmetic operators:

- Addition: `+`
- Subtraction: `-`
- Multiplication: `*`
- Division: `/`
- Exponentiation: `^`
- Modulo: `%%`

The last two might need some explaining:
- The `^` operator raises the number to its left to the power of the number to its right: for example `3^2` is 9.
- The modulo returns the remainder of the division of the number to the left by the number on its right, for example 5 modulo 3 or `5 %% 3` is 2.

*** =instructions
- Type `2^5` in the editor to calculate 2 to the power 5.
- Type `28 %% 6` to calculate 28 modulo 6.
- Click 'Submit Answer' and have a look at the R output in the console.
- Note how the `#` symbol is used to add comments on the R code.

*** =hint
Another example of the modulo operator: `9 %5 2` equals `1`.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# An addition
5 + 5 

# A subtraction
5 - 5 

# A multiplication
3 * 5

 # A division
(5 + 5)/2 

# Exponentiation


# Modulo

```

*** =solution
```{r eval=FALSE}
# An addition
5 + 5

# A subtraction
5 - 5 

# A multiplication
3 * 5

 # A division
(5 + 5)/2 

# Exponentiation
2^5

# Modulo
28 %% 6
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove the other arithmetic examples!"
test_output_contains("5 + 5", incorrect_msg = msg)
test_output_contains("5 - 5", incorrect_msg = msg)
test_output_contains("(5 + 5)/2", incorrect_msg = msg)
test_output_contains("2^5", incorrect_msg = "The exponentiation example is not correct. Make sure to read the instructions carefully.")
test_output_contains("28 %% 6", incorrect_msg = "There seems to be an issue with the modulo example. Make sure to read the instructions carefully.")
success_msg("Great! Head over to the next exercise.")
```

---

## Variable assignment 

A basic concept in (statistical) programming is called a **variable**. 

A variable allows you to store a value (e.g. 4) or an object (e.g. a function description) in R. You can then later use this variable's name to easily access the value or the object that is stored within this variable. 

You can assign a value 4 to a variable `my_variable` with the command

```
my_variable <- 4
```

*** =instructions
Over to you: complete the code in the editor such that it assigns the value 42 to the variable `x` in the editor. Click 'Submit Answer'. Notice that when you ask R to print `x`, the value 42 appears.

*** =hint
Look at how the value 4 was assigned to `my_variable` in the exercise's assignment. Do the exact same thing in the editor, but now assign 42 to the variable `x`.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Assign the value 42 to 'x'
x <- 

# Print out the value of the variable 'x'
x
```

*** =solution
```{r eval=FALSE}
# Assign the value 42 to 'x'
x <- 42

# Print out the value of the variable 'x'
x
```

*** =sct
```{r eval=FALSE}
test_object("x", undefined_msg = "Make sure to define a variable <code>x</code>.",
            incorrect_msg = "Make sure that you assign the correct value to <code>x</code>.") 
success_msg("Good job! Have you noticed that R does not print the value of a variable to the console when you did the assingment? <code>x <- 42</code> did not generate any output, because R assumes that you will be needing this variable in the future. Otherwise you wouldn't have stored the value in a variable in the first place, right? Proceed to the next exercise!")
```

---

## Variable assignment (2)

Suppose you have a fruit basket with five apples. As a data analyst in training, you want to store the number of apples in a variable with the name `my_apples`. 

*** =instructions
- Type the following code in the editor: `my_apples <- 5`. This will assign the value 5 to `my_apples`.
- Type: `my_apples` below the second comment. This will print out the value of `my_apples`.
- Click 'Submit Answer', and look at the console: you see that the number 5 is printed. So R now links the variable `my_apples` to the value 5.

*** =hint
Remember that if you want to assign a number or an object to a variable in R, you can make use of the assignment operator `<-`. Alternativy, you can use `=`, but `<-` is widely preferred in the R community.

*** =pre_exercise_code
```{r eval=FALSE}
```

*** =sample_code
```{r eval=FALSE}
# Assign the value 5 to the variable called 'my_apples'


# Print out the value of the variable 'my_apples'

```

*** =solution
```{r eval=FALSE}
# Assign the value 5 to the variable called 'my_apples'
my_apples <- 5

# Print out the value of the variable 'my_apples'
my_apples
```

*** =sct
```{r eval=FALSE}
test_object("my_apples", 
            undefined_msg = "Please make sure to define a variable <code>my_apples</code>.",
            incorrect_msg = "Make sure that you assign the correct value to <code>my_apples</code>.")
test_output_contains("my_apples", 
                     incorrect_msg = "Have you explicitly told R to print out the <code>my_apples</code> variable to the console?")
success_msg("Great! Continue to the next exercise!")
```

---

## Variable assignment (3)

Every tasty fruit basket needs oranges, so you decide to add six oranges. As a data analyst, your reflex is to immediately create the variable `my_oranges` and assign the value 6 to it. Next, you want to calculate how many pieces of fruit you have in total. Since you have given meaningful names to these values, you can now code this in a clear way: 
```
my_apples + my_oranges
```

*** =instructions
- Assign to `my_oranges` the value 6.
- R allows you to combine the variables `my_apples` and `my_oranges` into a new variable `my_fruit`. Create this new variable `my_fruit`, which is the total amount of fruits in your fruit basket.

*** =hint
`my_fruit` is just the sum of `my_apples` and `my_oranges`. You can use the `+` operator to sum the two and `<-` to assign that value to the variable `my_fruit`.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Assign a value to the variables called 'my_apples' and 'my_oranges'
my_apples <- 5


# Add these two variables together and print the result


# Create the variable 'my_fruit'

```

*** =solution
```{r eval=FALSE}
# Assign a value to the variables called 'my_apples' and 'my_oranges'
my_apples  <- 5
my_oranges <- 6

# Add these two variables together and print the result
my_apples + my_oranges

# Create the variable 'my_fruit'
my_fruit <- my_apples + my_oranges
```

*** =sct
```{r eval=FALSE}
test_object("my_apples", 
            undefined_msg = "Please make sure to define a variable <code>my_apples</code>.",
            incorrect_msg = "Do not change the assignment of the <code>my_apples</code> variable!")
test_object("my_oranges", 
            undefined_msg = "Have you defined a variable <code>my_oranges</code>?",
            incorrect_msg = "Have you assigned the correct value to the variable <code>my_oranges</code>?")
test_output_contains("my_apples + my_oranges",
                     incorrect_msg = "The output does not contain the result of adding <code>my_apples</code> and <code>my_oranges</code>. Try again")
test_object("my_fruit",
            undefined_msg = "Please make sure to define a variable <code>my_fruit</code>.",
            incorrect_msg = "Make sure that you assign the correct value to <code>my_fruit</code>.")
success_msg("Nice one! The great advantage of doing calculations with variables is reusability. If you just change <code>my_apples</code> to equal 12 instead of 5 and rerun the script, <code>my_fruit</code> will automatically update as well. Continue to the next exercise.")
```

---
## Apples and oranges

Common knowledge tells you not to add apples and oranges. But hey, that is what you just did, no :-)? The `my_apples` and `my_oranges` variables both contained a number in the previous exercise. The `+` operator works with numeric variables in R. If you really tried to add "apples" and "oranges", and assigned a text value to the variable `my_oranges` (see the editor), you would be trying to assign the addition of a numeric and a character variable to the variable `my_fruit`. This is not possible.

*** =instructions
- Click 'Submit Answer' and read the error message. Make sure to understand why this did not work.
- Adjust the code so that R knows you have 6 oranges and thus a fruit basket with 11 pieces of fruit.

*** =hint
You have to assign the numeric value `6` to the `my_oranges` variable instead of the character value `"six"`. Note how the quotation marks are used to indicate that `"six"` is a character.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Assign a value to the variable called 'my_apples'
my_apples <- 5 

# Print out the value of 'my_apples'
my_apples       

# Assign a value to the variable 'my_oranges' and print it out
my_oranges <- "six" 
my_oranges 

# New variable that contains the total amount of fruit
my_fruit <- my_apples + my_oranges 
my_fruit
```

*** =solution
```{r eval=FALSE}
# Assign a value to the variable called my_apples
my_apples <- 5  

# Print out the value of answer
my_apples  

# Assign a value to the variable 'my_oranges' and print it out
my_oranges <- 6
my_oranges 

# New variable that contains the total amount of fruit
my_fruit <- my_apples + my_oranges 
my_fruit
```

*** =sct
```{r eval=FALSE}
test_error()
test_object("my_oranges",
            undefined_msg = "Please make sure to define a variable <code>my_oranges</code>.",
            incorrect_msg = "Make sure that you assign the correct value to <code>my_oranges</code>. Strings and numbers are different things!")
test_output_contains("my_apples + my_oranges", 
                     incorrect_msg = "The output does not contain the result of adding <code>my_apples</code> and <code>my_oranges</code>.")
test_object("my_fruit", 
            undefined_msg = "Please make sure to define a variable <code>my_fruit</code>.",
            incorrect_msg = "Make sure that you assign the correct value to <code>my_fruit</code>.")
success_msg("Awesome, keep up the good work! Continue to the next exercise.")
```

---

## Basic data types in R

R works with numerous data types. Some of the most basic types to get started are:

- Decimals values like `4.5` are called **numerics**.
- Natural numbers like `4` are called **integers**. Integers are also numerics.
- Boolean values (`TRUE` or `FALSE`) are called **logical** (`TRUE` can be abbreviated to `T` and `FALSE` to `F`).
- Text (or string) values are called **characters**.

Note how the quotation marks on the right indicate that "some text" is a character.

*** =instructions
Change the value of the:
- `my_numeric` variable to `42`.
- `my_character` variable to `"forty-two"`. Note that the quotation marks indicate that `"forty-two"` is a character.
- `my_logical` variable to `FALSE`.

Note that R is case sensitive!

*** =hint 
Replace the values in the editor with the values that are provided in the exercise. For example: 
`my_numeric <- 42` assigns the value 42 to the variable `my_numeric`. 

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# What is the answer to the universe?
my_numeric <- 42.5

# The quotation marks indicate that the variable is of type character
my_character <- "some text"

my_logical <- TRUE
```

*** =solution
```{r eval=FALSE}
# What is the answer to the universe?
my_numeric <- 42

# The quotation marks indicate that the variable is of type character
my_character <- "forty-two"

my_logical <- FALSE
```

*** =sct
```{r eval=FALSE}
test_object("my_numeric", 
            undefined_msg = "Please make sure to define a variable <code>my_numeric</code>.",
            incorrect_msg = "Make sure that you assign the correct value to <code>my_numeric.</code>")
test_object("my_character",
            undefined_msg = "Please make sure to define a variable <code>my_character</code>.",
            incorrect_msg = "Make sure that you assign the correct value to <code>my_character</code>. Do not forget the quotes.")
test_object("my_logical",
            undefined_msg = "Please make sure to define a variable <code>my_logical</code>.",
            incorrect_msg = "Make sure that you assign the correct value to <code>my_logical</code>.") 
success_msg("Great work! Continue to the next exercise.")
```

---

## What's that data type?

Do you remember that when you added `5 + "six"`, you got an error due to a mismatch in data types? You can avoid such embarrassing situations by checking the data type of a variable beforehand. You can do this as follows:

```
class(some_variable_name)
```

*** =instructions
Complete the code in the editor and print the class of `my_numeric`, `my_character` and `my_logical` to the console. 

*** =hint
You can find out the data type of the `my_numeric` variable for example by typing `class(my_numeric)`. What you should do is similar for `my_character` and `my_logical`. 

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Declare variables of different types
my_numeric <- 42
my_character <- "forty-two"
my_logical <- FALSE 

# Check which type these variables have:

```

*** =solution
```{r eval=FALSE}
# Declare variables of different types:
my_numeric <- 42
my_character <- "forty-two"
my_logical <- FALSE

# Check which type these variables have:
class(my_numeric)
class(my_character)
class(my_logical)
```

*** =sct
```{r eval=FALSE}
test_object("my_numeric")
test_object("my_character")
test_object("my_logical")
test_output_contains("class(my_numeric)",
                     incorrect_msg = "The output does not contain the class of <code>my_numeric</code>. Try again.")
test_output_contains("class(my_character)",
                     incorrect_msg = "The output does not contain the class of <code>my_character</code>. Give it another try.")
test_output_contains("class(my_logical)",
                     incorrect_msg = "The output does not contain the class of <code>my_logical</code>.")
success_msg("Congratulations! This was the last exercise for this chapter. Head over to the next chapter and get immersed in the world of vectors!")
```
