---
title_meta: Chapter 2
title: Vectors
description: >-
  We take you on a trip to Vegas, where you will learn how to analyze your
  gambling results using vectors in R. After completing this chapter, you will
  be able to create vectors in R, name them, select elements from them, and
  compare different vectors.
---

## Create a vector

```yaml
type: NormalExercise
key: d9b453dbdd
xp: 100
skills:
  - 1
```

Feeling lucky? You better, because this chapter takes you on a trip to the City of Sins, also known as *Statisticians Paradise*!

Thanks to R and your new data-analytical skills, you will learn how to uplift your performance at the tables and fire off your career as a professional gambler. This chapter will show how you can easily keep track of your betting progress and how you can do some simple analyses on past actions. Next stop, Vegas Baby... VEGAS!!

`@instructions`
- Do you still remember what you have learned in the first chapter? Assign the value `"Go!"` to the variable `vegas`. Remember: R is case sensitive!

`@hint`
Just type the following line in the editor:
```
vegas <- "Go!"
```

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
# Define the variable vegas
vegas <- 
```

`@solution`
```{r}
# Define the variable vegas
vegas <- "Go!"
```

`@sct`
```{r}
ex() %>% check_object("vegas") %>% check_equal(incorrect_msg = "Make sure that you assign the correct value to `vegas`. Do not forget that R is case sensitive!")

success_msg("Great! Head over to the next exercise.")
```

---

## Create a vector (2)

```yaml
type: NormalExercise
key: fd427db76f
xp: 100
skills:
  - 1
```

Let us focus first! 

On your way from rags to riches, you will make extensive use of vectors. Vectors are one-dimension arrays that can hold numeric data, character data, or logical data. In other words, a vector is a simple tool to store data. For example, you can store your daily gains and losses in the casinos. 

In R, you create a vector with the combine function [`c()`](http://www.rdocumentation.org/packages/base/functions/c). You place the vector elements separated by a comma between the parentheses. For example:

```
numeric_vector <- c(1, 2, 3)
character_vector <- c("a", "b", "c")
```

Once you have created these vectors in R, you can use them to do calculations.

`@instructions`
Complete the code such that `boolean_vector` contains the three elements: `TRUE`, `FALSE` and `TRUE` (in that order).

`@hint`
Assign `c(TRUE, FALSE, TRUE)` to the variable `boolean_vector` with the `<-` operator.

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")

# Complete the code for boolean_vector
boolean_vector <-
```

`@solution`
```{r}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")

# Complete the code for boolean_vector
boolean_vector <- c(TRUE, FALSE, TRUE)
```

`@sct`
```{r}
msg <- "Do not change the code that defined `numeric_vector` and `character_vector`!"
ex() %>% check_object("numeric_vector", undefined_msg = msg) %>% check_equal(, incorrect_msg = msg)

ex() %>% check_object("character_vector", undefined_msg = msg)  %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("boolean_vector") %>% check_equal(incorrect_msg = "Make sure that you assign the correct values to `boolean_vector`. Use `c(TRUE, FALSE, TRUE)`. Don't place quotes around `TRUE` and `FALSE`! Also, make sure to adopt the same order as listed in the instructions.")


success_msg("Perfect! Notice that adding a space behind the commas in the `c()` function improves the readability of your code. Let's practice some more with vector creation in the next exercise.")
```

---

## Create a vector (3)

```yaml
type: NormalExercise
key: 9f41229dbc
xp: 100
skills:
  - 1
```

After one week in Las Vegas and still zero Ferraris in your garage, you decide that it is time to start using your data analytical superpowers.

Before doing a first analysis, you decide to first collect all the winnings and losses for the last week: 

For `poker_vector`: 

- On Monday you won $140
- Tuesday you lost $50
- Wednesday you won $20 
- Thursday you lost $120
- Friday you won $240

For `roulette_vector`: 

- On Monday you lost $24
- Tuesday you lost $50
- Wednesday you won $100
- Thursday you lost $350
- Friday you won $10

You only played poker and roulette, since there was a delegation of mediums that occupied the craps tables. To be able to use this data in R, you decide to create the variables `poker_vector` and `roulette_vector`.

`@instructions`
Assign the winnings/losses for roulette to the variable `roulette_vector`. You won $140, then lost $50, won $20, lost $120, and won $240.

`@hint`
To help you with this step, the editor already contains the code for creating `poker_vector`. Assign the correct values to `roulette_vector` based on the numbers in the assignment. Do not forget that losses are negative numbers.

`@pre_exercise_code`
```{r}

```

`@sample_code`
```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <-  
```

`@solution`
```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)
```

`@sct`
```{r}
ex() %>% check_object("poker_vector") %>% check_equal(incorrect_msg = "Make sure that you assign the correct values to `poker_vector`.")

ex() %>% check_object("roulette_vector") %>% check_equal(incorrect_msg = "Make sure that you assign the correct values to `roulette_vector`. Make sure to adopt the correct order!")

success_msg("Very good! To check out the contents of your vectors, remember that you can always simply type the variable in the console and hit Enter. Proceed to the next exercise!")
```

---

## Naming a vector

```yaml
type: NormalExercise
key: 3b0b80b192
xp: 100
skills:
  - 1
```

As a data analyst, it is important to have a clear view on the data that you are using. Understanding what each element refers to is therefore essential. 

In the previous exercise, we created a vector with your winnings over the week. Each vector element refers to a day of the week but it is hard to tell which element belongs to which day. It would be nice if you could show that in the vector itself. 

You can give a name to the elements of a vector with the `names()` function. Have a look at this example:

```
some_vector <- c("John Doe", "poker player")
names(some_vector) <- c("Name", "Profession")
```

This code first creates a vector `some_vector` and then gives the two elements a name. The first element is assigned the name `Name`, while the second element is labeled `Profession`. Printing the contents to the console yields following output:

```
          Name     Profession 
    "John Doe" "poker player" 
```

`@instructions`
- The code in the editor names the elements in `poker_vector` with the days of the week. Add code to do the same thing for `roulette_vector`.

`@hint`
You can use `names(roulette_vector)` to set the names of the variable `roulette_vector`. Make sure to use the same vector with the days of the week as names. Remember that R is case sensitive!

`@pre_exercise_code`
```{r}

```

`@sample_code`
```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Assign days as names of poker_vector
names(poker_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# Assign days as names of roulette_vector

```

`@solution`
```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Assign days as names of poker_vector
names(poker_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# Assign days as names of roulette_vector
names(roulette_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
```

`@sct`
```{r}
ex() %>% check_object("poker_vector") %>% check_equal(incorrect_msg = "Do not change the values inside `poker_vector`; they were already coded for you.")

ex() %>% check_object("roulette_vector") %>% check_equal(incorrect_msg = "Do not change the values inside `roulette_vector`; they were already coded for you.")

ex() %>% check_object("poker_vector") %>% check_equal(eq_condition = 'equal', incorrect_msg = "Do not change the code that names the elements in `poker_vector`; focus on `roulette_vector`!")

ex() %>% check_object("roulette_vector") %>% check_equal(eq_condition = 'equal',incorrect_msg = "Make sure that you assign the correct names vector to `roulette_vector`. Use the exact same vector as the one that was used to name `poker_vector`.")


success_msg("Well done! Continue to the next exercise.")
```

---

## Naming a vector (2)

```yaml
type: NormalExercise
key: 6858c65a4a
xp: 100
skills:
  - 1
```

If you want to become a good statistician, you have to become lazy. (If you are already lazy, chances are high you are one of those exceptional, natural-born statistical talents.)

In the previous exercises you probably experienced that it is boring and frustrating to type and retype information such as the days of the week. However, when you look at it from a higher perspective, there is a more efficient way to do this, namely, to assign the days of the week vector to a **variable**! 

Just like you did with your poker and roulette returns, you can also create a variable that contains the days of the week. This way you can use and re-use it.

`@instructions`
- A variable `days_vector` that contains the days of the week has already been created for you.
- Use `days_vector` to set the names of `poker_vector` and `roulette_vector`.

`@hint`
You can use `names(poker_vector) <- days_vector` to set the names of the elements `poker_vector`. Do a similar thing for `roulette_vector`.

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# The variable days_vector
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
 
# Assign the names of the day to roulette_vector and poker_vector
names(poker_vector) <-   
names(roulette_vector) <-
```

`@solution`
```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# The variable days_vector
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# Assign the names of the day to roulette_vector and poker_vector
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector
```

`@sct`
```{r}
msg <- "Do not changes the predefined variables `poker_vector`, `roulette_vector` or `days_vector`."

ex() %>% check_object("poker_vector", undefined_msg = msg) %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("roulette_vector", undefined_msg = msg) %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("days_vector", undefined_msg = msg)  %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("poker_vector") %>% check_equal(incorrect_msg = "Make sure that you assign `days_vector` to the names of `poker_vector`.", eq_condition = "equal",)

ex() %>% check_object("roulette_vector") %>% check_equal(eq_condition = "equal", incorrect_msg = "Make sure that you assign `days_vector` to the names of `roulette_vector`.")


success_msg("Nice one! A word of advice: try to avoid code duplication at all times. Continue to the next exercise and learn how to do arithmetic with vectors!")
```

---

## Calculating total winnings

```yaml
type: NormalExercise
key: da995f099f
xp: 100
skills:
  - 1
```

Now that you have the poker and roulette winnings nicely as named vectors, you can start doing some data analytical magic. 

You want to find out the following type of information:

- How much has been your overall profit or loss per day of the week?
- Have you lost money over the week in total?
- Are you winning/losing money on poker or on roulette?

To get the answers, you have to do arithmetic calculations on vectors. 

It is important to know that if you sum two vectors in R, it takes the element-wise sum. For example, the following three statements are completely equivalent:

```
c(1, 2, 3) + c(4, 5, 6)
c(1 + 4, 2 + 5, 3 + 6)
c(5, 7, 9)
```

You can also do the calculations with variables that represent vectors:

```
a <- c(1, 2, 3) 
b <- c(4, 5, 6)
c <- a + b
```

`@instructions`
- Take the sum of the variables `A_vector` and `B_vector` and assign it to `total_vector`.
- Inspect the result by printing out `total_vector`.

`@hint`
Use the `+` operator to sum `A_vector` and `B_vector`. Use `<-` to assign the result to `total_vector`.

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
A_vector <- c(1, 2, 3)
B_vector <- c(4, 5, 6)

# Take the sum of A_vector and B_vector
total_vector <- 
  
# Print out total_vector

```

`@solution`
```{r}
A_vector <- c(1, 2, 3)
B_vector <- c(4, 5, 6)

# Take the sum of A_vector and B_vector
total_vector <- A_vector + B_vector

# Print out total_vector
total_vector
```

`@sct`
```{r}
msg <- "Do not change the contents of `A_vector` or `B_vector`!"
ex() %>% check_object("A_vector", undefined_msg = msg) %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("B_vector", undefined_msg = msg) %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("total_vector") %>% check_equal(incorrect_msg = "Make sure that `total_vector` contains the sum of `A_vector` and `B_vector`.")

ex() %>% check_output_expr("total_vector", missing_msg = "Don't forget to print out `total_vector`! Simply write `total_vector` on a new line.")

success_msg("Good job! Continue to the next exercise.")
```

---

## Calculating total winnings (2)

```yaml
type: NormalExercise
key: 2969d8ed65
xp: 100
skills:
  - 1
```

Now you understand how R does arithmetic with vectors, it is time to get those Ferraris in your garage! First, you need to understand what the overall profit or loss per day of the week was. The total daily profit is the sum of the profit/loss you realized on poker per day, and the profit/loss you realized on roulette per day. 

In R, this is just the sum of `roulette_vector` and `poker_vector`.

`@instructions`
Assign to the variable `total_daily` how much you won or lost on each day in total (poker and roulette combined).

`@hint`
Similar to the previous exercise, assign the sum of two vectors to a new variable, `total_daily`.

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Assign to total_daily how much you won/lost on each day
total_daily <- 
```

`@solution`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Assign to total_daily how much you won/lost on each day
total_daily <- poker_vector + roulette_vector
```

`@sct`
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
ex() %>% check_object("days_vector", undefined_msg = msg) %>% check_equal( incorrect_msg = msg)

ex() %>% check_object("poker_vector",  undefined_msg = msg) %>% check_equal(eq_condition = "equal", incorrect_msg = msg)

ex() %>% check_object("roulette_vector", undefined_msg = msg) %>% check_equal(eq_condition = "equal", incorrect_msg = msg)

ex() %>% check_object("total_daily") %>% check_equal(incorrect_msg = "Make sure that you assign the sum of `poker_vector` and `roulette_vector` to `total_daily`.")


success_msg("Great! Continue to the next exercise.")
```

---

## Calculating total winnings (3)

```yaml
type: NormalExercise
key: e66a56b9f0
xp: 100
skills:
  - 1
```

Based on the previous analysis, it looks like you had a mix of good and bad days. This is not what your ego expected, and you wonder if there may be a very tiny chance you have lost money over the week in total? 

A function that helps you to answer this question is [`sum()`](http://www.rdocumentation.org/packages/base/functions/sum). It calculates the sum of all elements of a vector. For example, to calculate the total amount of money you have lost/won with poker you do: 

```
total_poker <- sum(poker_vector)
```

`@instructions`
- Calculate the total amount of money that you have won/lost with roulette and assign to the variable `total_roulette`.
- Now that you have the totals for roulette and poker, you can easily calculate `total_week` (which is the sum of all gains and losses of the week).
- Print out `total_week`.

`@hint`
Use the [`sum()`](http://www.rdocumentation.org/packages/base/functions/sum) function to get the total of the `roulette_vector`. `total_week` is then the sum of `roulette_vector` and `poker_vector`.

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Total winnings with poker
total_poker <- sum(poker_vector)

# Total winnings with roulette
total_roulette <-  

# Total winnings overall
total_week <- 

# Print out total_week
  
```

`@solution`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Total winnings with poker
total_poker <- sum(poker_vector)

# Total winnings with roulette
total_roulette <-  sum(roulette_vector)

# Total winnings overall
total_week <- total_roulette + total_poker

# Print out total_week
total_week
```

`@sct`
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
ex()  %>% check_object("days_vector", undefined_msg = msg)  %>% check_equal(incorrect_msg = msg)
ex() %>% check_object("poker_vector",  undefined_msg = msg, ) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("roulette_vector", undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)


ex() %>% check_object("total_poker") %>% check_equal(incorrect_msg = "Make sure that you assign to `total_poker` the sum of the `poker_vector`.")

ex() %>% check_object("total_roulette") %>% check_equal(incorrect_msg = "Make sure that you assign to `total_roulette` the sum of the `roulette_vector`.")

ex() %>% check_object("total_week") %>% check_equal(incorrect_msg = "Make sure that you assign to `total_week` the sum of the other two total vectors: `total_roulette` and `total_poker`.")

ex() %>% check_output_expr("total_week", missing_msg = "Don't forget to write `total_week` on a new line to print out the variable.")

success_msg("Well done. This is pretty bad news...")
```

---

## Comparing total winnings

```yaml
type: NormalExercise
key: f532f5332d
xp: 100
skills:
  - 1
```

Oops, it seems like you are losing money. Time to rethink and adapt your strategy! This will require some deeper analysis... 

After a short brainstorm in your hotel's jacuzzi, you realize that a possible explanation might be that your skills in roulette are not as well developed as your skills in poker. So maybe your total gains in poker are higher (or `>` ) than in roulette.

`@instructions`
- Calculate `total_poker` and `total_roulette` as in the previous exercise. Use the `sum()` function twice.
- Check if your total gains in poker are higher than for roulette by using a comparison. Simply print out the result of this comparison. What do you conclude, should you focus on roulette or on poker?

`@hint`
- You partly calculated the answer to this question in the previous exercise already!
- To check if 6 is larger than 5, you type `6 > 5`. This returns a logical value (`TRUE` or `FALSE`).

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Calculate total gains for poker and roulette
total_poker <-
total_roulette <-

# Check if you realized higher total gains in poker than in roulette 

```

`@solution`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Calculate total gains for poker and roulette
total_poker <- sum(poker_vector)
total_roulette <- sum(roulette_vector)

# Check if you realized higher total gains in poker than in roulette
total_poker > total_roulette
```

`@sct`
```{r}
msg <- "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
ex()  %>% check_object("days_vector", undefined_msg = msg)  %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("poker_vector", undefined_msg = msg)  %>% check_equal(eq_condition = "equal", incorrect_msg = msg)

ex() %>% check_object("roulette_vector",  undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)


ex()  %>% check_object("total_poker")  %>% check_equal(incorrect_msg = "Make sure that you assign to `total_poker` the sum of the `poker_vector`. Use `sum()`.")
ex()  %>% check_object("total_roulette")  %>% check_equal(incorrect_msg = "Make sure that you assign to `total_roulette` the sum of the `roulette_vector`. Use `sum()`.")
ex() %>% check_output_expr("total_poker > total_roulette",missing_msg = "Have you correctly carried out the comparison? To check if `total_poker` is greater than `total_roulette`, you can use `total_poker > total_roulette`.")

success_msg("Good job! Continue to the next exercise.")
```

---

## Vector selection: the good times

```yaml
type: NormalExercise
key: 8d78be44e9
xp: 100
skills:
  - 1
```

Your hunch seemed to be right. It appears that the poker game is more your cup of tea than roulette. 

Another possible route for investigation is your performance at the beginning of the working week compared to the end of it. You did have a couple of Margarita cocktails at the end of the week... 

To answer that question, you only want to focus on a selection of the `total_vector`. In other words, our goal is to select specific elements of the vector. To select elements of a vector (and later matrices, data frames, ...), you can use square brackets. Between the square brackets, you indicate what elements to select. For example, to select the first element of the vector, you type `poker_vector[1]`. To select the second element of the vector, you type `poker_vector[2]`, etc. Notice that the first element in a vector has index 1, not 0 as in many other programming languages.

`@instructions`
Assign the poker results of Wednesday to the variable `poker_wednesday`.

`@hint`
Wednesday is the third element of `poker_vector`, and can thus be selected with `poker_vector[3]`.

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Define a new variable based on a selection
poker_wednesday <- 
```

`@solution`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Define a new variable based on a selection
poker_wednesday <- poker_vector[3]
```

`@sct`
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
ex() %>% check_object("days_vector", undefined_msg = msg) %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("poker_vector",  undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("roulette_vector", undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("poker_wednesday", undefined_msg = "Please make sure to define a variable `poker_wednesday`.") %>% check_equal(incorrect_msg = "It looks like `poker_wednesday` does not contain the correct value of the `poker_vector`.")


success_msg("Great! R also makes it possible to select multiple elements from a vector at once. Learn how in the next exercise!")
```

---

## Vector selection: the good times (2)

```yaml
type: NormalExercise
key: '1351521670'
xp: 100
skills:
  - 1
```

How about analyzing your midweek results? 

To select multiple elements from a vector, you can add square brackets at the end of it. You can indicate between the brackets what elements should be selected. For example: suppose you want to select the first and the fifth day of the week: use the vector `c(1, 5)` between the square brackets. For example, the code below selects the first and fifth element of `poker_vector`:

```
poker_vector[c(1, 5)]
```

`@instructions`
Assign the poker results of Tuesday, Wednesday and Thursday to the variable `poker_midweek`.

`@hint`
Use the vector `c(2, 3, 4)` between square brackets to select the correct elements of `poker_vector`.

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Define a new variable based on a selection
poker_midweek <- 
```

`@solution`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Define a new variable based on a selection
poker_midweek <- poker_vector[c(2, 3, 4)]
```

`@sct`
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
ex() %>% check_object("days_vector", undefined_msg = msg) %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("poker_vector", undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("roulette_vector",  undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("poker_midweek") %>% check_equal(incorrect_msg = "It looks like `poker_midweek` does not contain the correct values from `poker_vector`. You can use the vector `c(2, 3, 4)` inside square brackets.")

success_msg("Well done! Continue to the next exercise to specialize in vector selection some more!");
```

---

## Vector selection: the good times (3)

```yaml
type: NormalExercise
key: 27976b79f4
xp: 100
skills:
  - 1
```

Selecting multiple elements of `poker_vector` with `c(2, 3, 4)` is not very convenient. Many statisticians are lazy people by nature, so they created an easier way to do this: `c(2, 3, 4)` can be abbreviated to`2:4`, which generates a vector with all natural numbers from 2 up to 4.

So, another way to find the mid-week results is `poker_vector[2:4]`. Notice how the vector `2:4` is placed between the square brackets to select element 2 up to 4.

`@instructions`
Assign to `roulette_selection_vector` the roulette results from Tuesday up to Friday; make use of `:` if it makes things easier for you.

`@hint`
Assign a selection of `roulette_vector` to `roulette_selection_vector` by placing `2:5` between square brackets.

`@pre_exercise_code`
```{r}

```

`@sample_code`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Define a new variable based on a selection
roulette_selection_vector <- 
```

`@solution`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Define a new variable based on a selection
roulette_selection_vector <- roulette_vector[2:5]
```

`@sct`
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
ex() %>% check_object("days_vector", undefined_msg = msg) %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("poker_vector",  undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("roulette_vector", undefined_msg = msg) %>% check_equal( eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("roulette_selection_vector", undefined_msg = "Please make sure to define a variable `roulette_selection_vector`.") %>% check_equal(incorrect_msg = "It looks like `roulette_selection_vector` does not contain the correct selection from `roulette_vector`. Make sure to to use the right indexes.")

success_msg("Awesome! The colon operator is extremely useful and very often used in R programming, so remember it well. Proceed to the next exercise.")
```

---

## Vector selection: the good times (4)

```yaml
type: NormalExercise
key: e6c263ddee
xp: 100
skills:
  - 1
```

Another way to tackle the previous exercise is by using the names of the vector elements (Monday, Tuesday, ...) instead of their numeric positions. For example, 

```
poker_vector["Monday"]
```

will select the first element of `poker_vector` since `"Monday"` is the name of that first element.

Just like you did in the previous exercise with numerics, you can also use the element names to select multiple elements, for example: 

```
poker_vector[c("Monday","Tuesday")]
```

`@instructions`
- Select the first three elements in `poker_vector` by using their names: `"Monday"`, `"Tuesday"` and `"Wednesday"`. Assign the result of the selection to `poker_start`.
- Calculate the average of the values in `poker_start` with the [`mean()`](http://www.rdocumentation.org/packages/base/functions/mean) function. Simply print out the result so you can inspect it.

`@hint`
- You can use `c("Monday", "Tuesday", "Wednesday")` inside square brackets to subset `poker_vector` appropriately.
- You can use `mean(poker_start)` to get the mean of the elements in `poker_start`. You do not need the mean of all poker elements, but only of the first three days.

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Select poker results for Monday, Tuesday and Wednesday
poker_start <- 
  
# Calculate the average of the elements in poker_start

```

`@solution`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Select poker results for Monday, Tuesday and Wednesday
poker_start <- poker_vector[c("Monday", "Tuesday", "Wednesday")]
  
# Calculate the average of the elements in poker_start
mean(poker_start)
```

`@sct`
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
ex() %>% check_object("days_vector", undefined_msg = msg) %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("poker_vector", undefined_msg = msg) %>% check_equal(eq_condition = "equal", incorrect_msg = msg)

ex() %>% check_object("roulette_vector", undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("poker_start") %>% check_equal(incorrect_msg = "It looks like `poker_start` does not contain the first three values of `poker_vector`. You can use `c(\"Monday\", \"Tuesday\", \"Wednesday\")` inside square brackets to do this.")

ex() %>% check_output_expr("mean(poker_start)", missing_msg = "Have you correctly calculated the average of the values in `poker_start` and printed it out? Use `mean(poker_start)`.")

success_msg("Good job! Apart from subsetting vectors by index or by name, you can also subset vectors by comparison. The next exercises will show you how!")
```

---

## Selection by comparison - Step 1

```yaml
type: NormalExercise
key: f0f619c901
xp: 100
skills:
  - 1
```

By making use of comparison operators, we can approach the previous question in a more proactive way. 

The (logical) comparison operators known to R are:

- `<` for less than
- `>` for greater than
- `<=` for less than or equal to
- `>=` for greater than or equal to
- `==` for equal to each other
- `!=` not equal to each other

As seen in the previous chapter, stating `6 > 5` returns `TRUE`. The nice thing about R is that you can use these comparison operators also on vectors. For example:

```
c(4, 5, 6) > 5
[1] FALSE FALSE TRUE
```

This command tests for every element of the vector if the condition stated by the comparison operator is `TRUE` or `FALSE`.

`@instructions`
- Check which elements in `poker_vector` are positive (i.e. > 0) and assign this to `selection_vector`. 
- Print out `selection_vector` so you can inspect it. The printout tells you whether you won (`TRUE`) or lost (`FALSE`) any money for each day.

`@hint`
In order to check for which days your poker gains are positive, R should check for each element of `poker_vector` whether it is larger than zero. `some_vector > 0` is the way to tell R what you are after.

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on poker?
selection_vector <- 
  
# Print out selection_vector

```

`@solution`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on poker?
selection_vector <- poker_vector > 0
  
# Print out selection_vector
selection_vector
```

`@sct`
```{r}
msg <- "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
ex() %>% check_object("days_vector", undefined_msg = msg) %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("poker_vector",  undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("roulette_vector",  undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("selection_vector") %>% check_equal(incorrect_msg = "It looks like `selection_vector` does not contain the correct result. Remember that R uses element wise operations for vectors.")

ex() %>% check_output_expr("selection_vector", missing_msg = "Don't forget to print out `selection_vector` by writing the variable name on a new line.")

success_msg("Great!")
```

---

## Selection by comparison - Step 2

```yaml
type: NormalExercise
key: 2754fc5cd4
xp: 100
skills:
  - 1
```

Working with comparisons will make your data analytical life easier. Instead of selecting a subset of days to investigate yourself (like before), you can simply ask R to return only those days where you realized a positive return for poker. 

In the previous exercises you used `selection_vector <- poker_vector > 0` to find the days on which you had a positive poker return. Now, you would like to know not only the days on which you won, but also how much you won on those days. 

You can select the desired elements, by putting `selection_vector` between the square brackets that follow `poker_vector`:

```
poker_vector[selection_vector]
```

R knows what to do when you pass a logical vector in square brackets: it will only select the elements that correspond to `TRUE` in `selection_vector`.

`@instructions`
Use `selection_vector` in square brackets to assign the amounts that you won on the profitable days to the variable `poker_winning_days`.

`@hint`
Use `poker_vector[selection_vector]` to select the desired elements from `poker_vector`, and assign the result to `poker_winning_days`.

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on poker?
selection_vector <- poker_vector > 0

# Select from poker_vector these days
poker_winning_days <- 
```

`@solution`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on poker?
selection_vector <- poker_vector > 0

# Select from poker_vector these days
poker_winning_days <- poker_vector[selection_vector]
```

`@sct`
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
ex() %>% check_object("days_vector", undefined_msg = msg) %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("poker_vector",  undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("roulette_vector",  undefined_msg = msg) %>% check_equal(eq_condition = "equal",incorrect_msg = msg)

ex() %>% check_object("selection_vector") %>% check_equal(incorrect_msg = "Don't change the way `selection_vector` is calculated.")

ex() %>% check_object("poker_winning_days") %>% check_equal(incorrect_msg = "It looks like `poker_winning_days` does not contain the correct result. Use `poker_vector[selection_vector]`.")

success_msg("Good job! Continue to the next exercise.")
```

---

## Advanced selection

```yaml
type: NormalExercise
key: 59e8dcbbd5
xp: 100
skills:
  - 1
```

Just like you did for poker, you also want to know those days where you realized a positive return for roulette.

`@instructions`
- Create the variable `selection_vector`, this time to see if you made profit with roulette for different days.
- Assign the amounts that you made on the days that you ended positively for roulette to the variable `roulette_winning_days`. This vector thus contains the positive winnings of `roulette_vector`.

`@hint`
Once you've correctly calculated `selection_vector`, you can again use `roulette_vector[selection_vector]` to select the positive results from `roulette_vector`.

`@pre_exercise_code`
```{r}
# no pec
```

`@sample_code`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on roulette?
selection_vector <-

# Select from roulette_vector these days
roulette_winning_days <- 
```

`@solution`
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)
roulette_vector <- c(-24, -50, 100, -350, 10)
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Which days did you make money on roulette?
selection_vector <- roulette_vector > 0

# Select from roulette_vector these days
roulette_winning_days <- roulette_vector[selection_vector]
```

`@sct`
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."

ex() %>% check_object("days_vector", undefined_msg = msg) %>% check_equal(incorrect_msg = msg)

ex() %>% check_object("poker_vector", undefined_msg = msg) %>% check_equal(eq_condition = "equal", incorrect_msg = msg)

ex() %>% check_object("roulette_vector", undefined_msg = msg) %>% check_equal(eq_condition = "equal", incorrect_msg = msg)

ex() %>% check_object("selection_vector") %>% check_equal(incorrect_msg = "It looks like `selection_vector` does not contain the correct result. Use `roulette_vector > 0`.")

ex() %>% check_object("roulette_winning_days") %>% check_equal(incorrect_msg = "It looks like `roulette_winning_days` does not contain the correct result. Use `roulette_vector[selection_vector]`.")


success_msg("Great! This exercise concludes the chapter on vectors. The next chapter will introduce you to the two-dimensional version of vectors: matrices.")
```
