--- 
title_meta  : Chapter 2
title       : Vectors
description : "In this free R course, we'll take you on a trip to Vegas, where you will learn how to analyze your gambling results using vectors in R! After completing this chapter, you will be able to create vectors in R, name them, select elements from them and compare different vectors."

--- type:NormalExercise xp:100 skills:1 key:d9b453dbdd
## Create a vector

Feeling lucky? You better, because this chapter takes you on a trip to the City of Sins, also known as *Statisticians Paradise*!

Thanks to R and your new data-analytical skills, you will learn how to uplift your performance at the tables and fire off your career as a professional gambler. This chapter will show how you can easily keep track of your betting progress and how you can do some simple analyses on past actions. Next stop, Vegas Baby... VEGAS!!

*** =instructions
- Do you still remember what you have learned in the first chapter? Assign the value `"Go!"` to the variable `vegas`. Remember: R is case sensitive!

*** =hint
Just type the following line in the editor:
```
vegas <- "Go!"
```

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Define the variable vegas
vegas <- 
```

*** =solution
```{r}
# Define the variable vegas
vegas <- "Go!"
```

*** =sct
```{r}
test_object("vegas", incorrect_msg = "Make sure that you assign the correct value to `vegas`. Do not forget that R is case sensitive!")
success_msg("Great! Head over to the next exercise.")
```


--- type:NormalExercise xp:100 skills:1 key:fd427db76f
## Create a vector (2)

Let us focus first! 

On your way from rags to riches, you will make extensive use of vectors. Vectors are one-dimension arrays that can hold numeric data, character data, or logical data. In other words, a vector is a simple tool to store data. For example, you can store your daily gains and losses in the casinos. 

In R, the [`c()`](http://www.rdocumentation.org/packages/base/functions/c) function is used to *c*ombine values to *c*reate vectors. You place the vector elements separated by a comma between the parentheses. For example:

```
numeric_vector <- c(1, 2, 3)
character_vector <- c("a", "b", "c")
```

Once you have created these vectors in R, you can use them to do calculations.

*** =instructions 

Complete the code by assigning a value to the variable named `logical_vector`.

* Call the [`c()`](http://www.rdocumentation.org/packages/base/functions/c) function with values `TRUE`, `FALSE` and `TRUE` (in that order).

*** =hint 
Assign `c(TRUE, FALSE, TRUE)` to the variable `logical_vector` with the `<-` operator.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")

# Complete the code for logical_vector
logical_vector <-
```

*** =solution
```{r}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")

# Complete the code for logical_vector
logical_vector <- c(TRUE, FALSE, TRUE)
```

*** =sct
```{r}
msg <- "Do not change the code that defined `numeric_vector` and `character_vector`!"
test_object("numeric_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("character_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("logical_vector",
            incorrect_msg = "Make sure that you assign the correct values to `logical_vector`. Use `c(TRUE, FALSE, TRUE)`. Don't place quotes around `TRUE` and `FALSE`! Also, make sure to adopt the same order as listed in the instructions.")
test_function("c", index = 3)
success_msg("Perfect! Notice that adding a space behind the commas in the `c()` function improves the readability of your code. Let's practice some more with vector creation in the next exercise.")
```


--- type:NormalExercise xp:100 skills:1 key:9f41229dbc
## Create a vector (3)

After one week in Las Vegas and still zero Ferraris in your garage, you decide that it is time to start using your data analytical superpowers.

Before doing a first analysis, you decide to first collect all the winnings and losses for the last week: 

For `poker_vector`: 

- On Monday you won $140.25
- Tuesday you lost $50.50
- Wednesday you won $20.25
- Thursday you lost $120.00
- Friday you won $240.00

For `roulette_vector`: 

- On Monday you lost $24.75
- Tuesday you lost $50.25
- Wednesday you won $100
- Thursday you lost $350.50
- Friday you won $10

To be able to use this data in R, you decide to create the variables `poker_vector` and `roulette_vector`.

*** =instructions

Assign the winnings/losses for roulette to the variable `roulette_vector`. Notice that for whole dollar values, you don't need to specify the number of cents.

*** =hint
To help you with this step, the editor already contains the code for creating `poker_vector`. Assign the correct values to `roulette_vector` based on the numbers in the assignment. Do not forget that losses are negative numbers.

*** =sample_code
```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140.25, -50.50, 20.25, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <-  
```

*** =solution
```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140.25, -50.50, 20.25, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24.75, -50.25, 100, -350.50, 10)
```

*** =sct
```{r}
test_object("poker_vector",
            incorrect_msg = "Make sure that you assign the correct values to `poker_vector`.")
test_object("roulette_vector",
            incorrect_msg = "Make sure that you assign the correct values to `roulette_vector`. Make sure to adopt the correct order!")
success_msg("Very good! To check out the contents of your vectors, remember that you can always simply type the variable in the console and hit Enter. Proceed to the next exercise!")
```


--- type:NormalExercise xp:100 skills:1 key:3b0b80b192
## Naming a vector

As a data analyst, it is important to have a clear view on the data that you are using. Understanding what each element refers to is therefore essential. 

In the previous exercise, we created a vector with your winnings over the week. Each vector element refers to a day of the week but it is hard to tell which element belongs to which day. It would be nice if you could show that in the vector itself. 

You can name elements of a vector using a `name = value` syntax. The names don't need to be quoted unless they include spaces or punctuation other than a `.` or `_`.  For example, 

```{r}
casinos <- c(Luxor = 1, Stratosphere = 2, "Caesars Palace" = 3)
```

You can also name the elements of a vector after it has been created with the [`names()`](https://www.rdocumentation.org/packages/base/topics/names) function. Have a look at this example:

```{r}
casinos <- c(1, 2, 3)
names(casinos) <- c("Luxor", "Stratosphere", "Caesars Palace")
```

In both cases, the value assigned to `casinos` is the same. You can use whichever method is most convenient for you. When you print the variable to the console, the names are included. The output is shown in comments.  

```
casinos
##          Luxor   Stratosphere Caesars Palace 
##              1              2              3
```

*** =instructions
- Complete the names of `poker_vector` with the remaining days of the week (`Thursday` and `Friday`), using `name = value` syntax. 
- Call the [`names()`](https://www.rdocumentation.org/packages/base/topics/names) function to complete the names of `roulette_vector` to the remaining days of the week.

*** =hint
You can use `names(roulette_vector)` to set the names of the variable `roulette_vector`. Make sure to use the same vector with the days of the week as names. Remember that R is case sensitive!

*** =sample_code
```{r}
# Poker winnings from Monday to Friday; complete the names
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, ___ = -120, ___ = 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24.75, -50.25, 100, -350.50, 10)

# Assign days as names of roulette_vector
___(roulette_vector) <- c("Monday", ___, ___, "Thursday", "Friday")
```

*** =solution
```{r}
# Poker winnings from Monday to Friday; complete the names
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24.75, -50.25, 100, -350.50, 10)

# Assign days as names of roulette_vector
names(roulette_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
```

*** =sct
```{r}
test_object("poker_vector",
            incorrect_msg = "Do not change the values inside `poker_vector`; they were already coded for you.")
test_object("roulette_vector",
            incorrect_msg = "Do not change the values inside `roulette_vector`; they were already coded for you.")
test_object("poker_vector",
            eq_condition = "equal",
            incorrect_msg = "Do not change the code that names the elements in `poker_vector`; focus on `roulette_vector`!")
test_object("roulette_vector",
            eq_condition = "equal",
            incorrect_msg = "Make sure that you assign the correct names vector to `roulette_vector`. Use the exact same vector as the one that was used to name `poker_vector`.")
success_msg("Well done! Continue to the next exercise.")
```


--- type:NormalExercise xp:100 skills:1 key:6858c65a4a
## Naming a vector (2)

If you want to become a good statistician, you have to become lazy. (If you are already lazy, the chances are high that you are one of those exceptional, natural-born statistical talents.)

In the previous exercises you probably experienced that it is boring and frustrating to type and retype information such as the days of the week. However, when you look at it from a higher perspective, there is a more efficient way to do this, namely, to assign the days of the week vector to a **variable**! 

Just like you did with your poker and roulette returns, you can also create a variable that contains the days of the week. This way you can use and re-use it.

*** =instructions
- A variable `days_vector` that contains the days of the week has already been created for you.
- Use `days_vector` to set the [`names()`](https://www.rdocumentation.org/packages/base/topics/names) of `poker_vector` and `roulette_vector`.

*** =hint
You can use `names(poker_vector) <- days_vector` to set the names of the elements `poker_vector`. Do a similar thing for `roulette_vector`.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140.25, -50.50, 20.25, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24.75, -50.25, 100, -350.50, 10)

# The variable days_vector
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
 
# Assign the names of the day to roulette_vector and poker_vector
___(poker_vector) <-   
___(roulette_vector) <-
```

*** =solution
```{r}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# The variable days_vector
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

# Assign the names of the day to roulette_vector and poker_vector
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector
```

*** =sct
```{r}
msg <- "Do not changes the pre-defined variables `poker_vector`, `roulette_vector` or `days_vector`."
test_object("poker_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("days_vector", undefined_msg = msg, incorrect_msg = msg)

test_object("poker_vector",
            eq_condition = "equal",
            incorrect_msg = "Make sure that you assign `days_vector` to the names of `poker_vector`.")
test_object("roulette_vector",
            eq_condition = "equal",
            incorrect_msg = "Make sure that you assign `days_vector` to the names of `roulette_vector`.")
success_msg("Nice one! A word of advice: try to avoid code duplication at all times. Continue to the next exercise and learn how to do arithmetic with vectors!")
```


--- type:NormalExercise xp:100 skills:1 key:da995f099f
## Calculating total winnings

Now that you have the poker and roulette winnings nicely as named vectors, you can start doing some data analytical magic. 

Suppose you want to calculate your total winnings for each day of the week. You could calculate the total winnings for Monday, then for Tuesday, and so on throughout the week. That sounds like a lot of effort, and as you learned in the last exercise, laziness is an important skill!

Fortunately, R makes it easy to do arithmetic calculations on vectors. If you sum two vectors in R, it takes the element-wise sum. For example, the following three statements are completely equivalent:

```
c(1, 2, 3) + c(4, 5, 6)
c(1 + 4, 2 + 5, 3 + 6)
c(5, 7, 9)
```

You can also do the calculations with variables that represent vectors:

```
first <- c(1, 2, 3) 
second <- c(4, 5, 6)
total <- a + b
```

This means that to do all your calculations, you only need one line of code.

*** =instructions
- Take the sum of the variables `square_numbers` and `triangular_numbers` and assign it to `total_vector`.
- Inspect the result by printing out `total_vector`.

*** =hint
Use the `+` operator to sum `square_numbers` and `triangular_numbers`. Use `<-` to assign the result to `total_vector`.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
square_numbers <- c(1, 4, 9, 16, 25, 36)
triangular_numbers <- c(1, 3, 6, 10, 15, 21)

# Take the sum of square_numbers and triangular_numbers
total_vector <- 
  
# Print out total_vector

```

*** =solution
```{r}
square_numbers <- c(1, 2, 3)
triangular_numbers <- c(4, 5, 6)

# Take the sum of square_numbers and triangular_numbers
total_vector <- square_numbers + triangular_numbers

# Print out total_vector
total_vector
```

*** =sct
```{r}
msg <- "Do not change the contents of `square_numbers` or `triangular_numbers`!"
test_object("square_numbers", undefined_msg = msg, incorrect_msg = msg)
test_object("triangular_numbers", undefined_msg = msg, incorrect_msg = msg)
test_object("total_vector", incorrect_msg = "Make sure that `total_vector` contains the sum of `square_numbers` and `triangular_numbers`.")
test_output_contains("total_vector", incorrect_msg = "Don't forget to print out `total_vector`! Simply write `total_vector` on a new line.")
success_msg("Good job! Continue to the next exercise.")
```


--- type:NormalExercise xp:100 skills:1 key:2969d8ed65
## Calculating total winnings (2)

Now you understand how R does arithmetic with vectors, it is time to get those Ferraris in your garage! First, you need to understand what the overall winnings per day of the week were. The total daily winnings are the sum of the winnings you realized on poker per day, and the winnings you realized on roulette per day. 

In R, this is just `roulette_vector` plus `poker_vector`.

*** =instructions

Assign to the variable `daily_winnings` how much you won or lost on each day in total (poker and roulette combined).

*** =hint
Similar to the previous exercise, assign the sum of two vectors to a new variable, `daily_winnings`.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Poker and roulette winnings from Monday to Friday
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Assign to daily_winnings how much you won/lost on each day
daily_winnings <- 
```

*** =solution
```{r}
# Poker and roulette winnings from Monday to Friday
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Assign to total_daily how much you won/lost on each day
daily_winnings <- poker_vector + roulette_vector
```

*** =sct
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)

test_object("daily_winnings", incorrect_msg = "Make sure that you assign the sum of `poker_vector` and `roulette_vector` to `daily_winnings`.")

success_msg("Yippee! Notice how the names of the inputs were copied to the total.")
```


--- type:NormalExercise xp:100 skills:1 key:e66a56b9f0
## Calculating total winnings (3)

Based on the previous analysis, it looks like you had a mix of good and bad days. This is not what your ego expected, and you wonder if there may be a very tiny chance you have lost money over the week in total? 

A function that helps you to answer this question is [`sum()`](http://www.rdocumentation.org/packages/base/functions/sum). It calculates the sum of all elements of a vector. For example, to calculate the sum of the first few Fibonacci numbers, just pass the vector of numbers to `sum()`.

```{r}
fibonacci_numbers <- c(1, 1, 2, 3, 5, 8, 13, 21)
total_numbers <- sum(fibonacci_numbers)
```

`sum()` also lets you pass in several vectors. In this case, it will calculate the sum of all the elements in all the vectors that you pass in.

```{r}
fibonacci_numbers <- c(1, 1, 2, 3, 5, 8, 13, 21)
motzkin_numbers <- c(1, 1, 2, 4, 9, 21, 51, 127)
total_numbers <- sum(fibonacci_numbers, motzkin_numbers)
```

*** =instructions

The code to calculate the total winnings over thewhole week for poker is shown in the script.

- Calculate the total amount of money that you have won/lost with roulette and assign to the variable `total_roulette`.
- Now that you have the totals for roulette and poker, you can easily calculate `total_week` (which is the sum of all gains and losses of the week).
- Print out `total_week`.

*** =hint
Use the [`sum()`](http://www.rdocumentation.org/packages/base/functions/sum) function to get the total of the `roulette_vector`. `total_week` is then the sum of `roulette_vector` and `poker_vector`.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Poker and roulette winnings from Monday to Friday
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Total winnings with poker
total_poker <- sum(poker_vector)

# Total winnings with roulette
total_roulette <-  

# Total winnings overall
total_week <- 

# Print out total_week
  
```

*** =solution
```{r}
# Poker and roulette winnings from Monday to Friday
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Total winnings with poker
total_poker <- sum(poker_vector)

# Total winnings with roulette
total_roulette <-  sum(roulette_vector)

# Total winnings overall
total_week <- sum(poker_vector, roulette_vector)

# Print out total_week
total_week
```

*** =sct
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("total_poker", 
            incorrect_msg = "Make sure that you assign to `total_poker` the sum of the `poker_vector`.")
test_object("total_roulette",
            incorrect_msg = "Make sure that you assign to `total_roulette` the sum of the `roulette_vector`.")
test_object("total_week",
            incorrect_msg = "Make sure that you assign to `total_week` the sum of the other two total vectors: `total_roulette` and `total_poker`.")

test_output_contains("total_week", incorrect_msg = "Don't forget to write `total_week` on a new line to print out the variable.")
success_msg("Well done. This is pretty bad news...")
```


--- type:NormalExercise xp:100 skills:1 key:f532f5332d
## Comparing total winnings

Oops, it seems like you are losing money. Time to rethink and adapt your strategy! This will require some deeper analysis... 

After a short brainstorm in your hotel's jacuzzi, you realize that a possible explanation might be that your skills in roulette are not as well developed as your skills in poker. This makes you wonder if your total gains in poker are greater than in roulette.

R has six ways of comparing numbers, using **comparison operators**. Just like the arithmetic operators (`+`) that you saw previously, they operate elementwise on vectors.

`x > y` returns `TRUE` whenever the element of `x` is **greater than** the corresponding element of `y`. Similarly

- `x < y` means `x` is **less than** `y`.
- `x >= y` means `x` is **greater than or equal to** `y`.
- `x >= y` means `x` is **less than or equal to** `y`.
- `x == y` means `x` is **equal to** `y`.
- `x != y` means `x` is **not equal to** `y`.

Here is an example of each of the six situations, with the result shown in comments.

```{r}
x <- c(0, 3, 6)
y <- c(1, 3, 5)
x > y
## [1] FALSE FALSE  TRUE
x < y
## [1]  TRUE FALSE FALSE
x >= y
## [1] FALSE  TRUE  TRUE
x <= y
## [1]  TRUE  TRUE FALSE
x == y
## [1] FALSE  TRUE FALSE
x != y
## [1]  TRUE FALSE  TRUE
```

*** =instructions
- Calculate `total_poker` and `total_roulette` as in the previous exercise. Use the `sum()` function twice.
- Check if your total gains in poker are higher than for roulette by using a comparison. Simply print out the result of this comparison. What do you conclude, should you focus on roulette or on poker?

*** =hint
- You partly calculated the answer to this question in the previous exercise already!
- To check if 6 is larger than 5, you type `6 > 5`. This returns a logical value (`TRUE` or `FALSE`).

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Calculate total gains for poker and roulette
total_poker <-
total_roulette <-

# Check if you realized higher total gains in poker than in roulette 

```

*** =solution
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Calculate total gains for poker and roulette
total_poker <- sum(poker_vector)
total_roulette <- sum(roulette_vector)

# Check if you realized higher total gains in poker than in roulette
total_poker > total_roulette
```

*** =sct
```{r}
msg <- "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)

test_object("total_poker", 
            incorrect_msg = "Make sure that you assign to `total_poker` the sum of the `poker_vector`. Use `sum()`.")
test_object("total_roulette",
            incorrect_msg = "Make sure that you assign to `total_roulette` the sum of the `roulette_vector`. Use `sum()`.")
test_output_contains("total_poker > total_roulette",
                     incorrect_msg = "Have you correctly carried out the comparison? To check if `total_poker` is greater than `total_roulette`, you can use `total_poker > total_roulette`.")
success_msg("Good job! Comparison operators work on each element of the vectors, just like arithmetic operators.")
```


--- type:NormalExercise xp:100 skills:1 key:8d78be44e9
## Vector selection: the good times

Your hunch seemed to be right. It appears that the poker game is more your cup of tea than roulette. 

Another possible route for investigation is your performance at the beginning of the working week compared to the end of it. You did have a couple of Margarita cocktails at the end of the week... 

To answer that question, you only want to focus on a selection of the `total_vector`. In other words, our goal is to select specific elements of the vector. To select elements of a vector (and later matrices, data frames, ...), you can use **square brackets**. Between the square brackets, you indicate what elements to select by passing a **positive integer** denoting the position of the element that you want to select. For example, to select the first element of the vector, you type `poker_vector[1]`. To select the second element of the vector, you type `poker_vector[2]`, etc. Notice that the first element in a vector has index 1, not 0 as in many other programming languages.

*** =instructions
Assign the poker results of Wednesday to the variable `poker_wednesday`.

*** =hint
Wednesday is the third element of `poker_vector`, and can thus be selected with `poker_vector[3]`.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Poker and roulette winnings from Monday to Friday
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Define a new variable based on a selection
poker_wednesday <- 
```

*** =solution
```{r}
# Poker and roulette winnings from Monday to Friday
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Define a new variable based on a selection
poker_wednesday <- poker_vector[3]
```

*** =sct
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_wednesday", 
            undefined_msg = "Please make sure to define a variable `poker_wednesday`.",
            incorrect_msg = "It looks like `poker_wednesday` does not contain the correct value of the `poker_vector`.")
success_msg("Great! R also makes it possible to select multiple elements from a vector at once. Learn how in the next exercise!")
```


--- type:NormalExercise xp:100 skills:1 key:1351521670
## Vector selection: the good times (2)

How about analyzing several days at once? 

Indexing by position also lets you select multiple elements from the vector.  To do this, simply place a vector of positive whole numbers inside the square brackets.  All the values should be between one and the length of the vector. For example: suppose you want to select the first and the fifth day of the week: use the vector `c(1, 5)` between the square brackets. The code below selects the first and fifth element of `poker_vector`. That is, the results from Monday and Friday.

```{r}
poker_vector[c(1, 5)]
```

For consecutive elements, there is a shortcut: you can write `a:b` rather than `c(a, a + 1, a + 2, ..., b)`. For example, these two statements are the same. (The output is shown in comments.)

```{r}
2:7
## [1] 2 3 4 5 6 7
c(2, 3, 4, 5, 6, 7)
## [1] 2 3 4 5 6 7
```

*** =instructions

* Create a sequence of numbers from the position of Tuesday (2) to Thursday (4), and assign the result to `index_midweek`.
* Use the index variable to select the poker results of Tuesday, Wednesday and Thursday, and assign the result to `poker_midweek`.

*** =hint

* The index can be created with `c(2, 3, 4)`, or its shortcut using the colon operator, `2:4`.
* Use square brackets, `[`, to select the correct elements of `poker_vector`.

*** =pre_exercise_code
```{r}
# no pec
``` 

*** =sample_code
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Create a sequence from 2 to 4
index_midweek <- 

# Define a new variable based on a selection
poker_midweek <- 
```

*** =solution
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Create a sequence from 2 to 4
index_midweek <- 2:4

# Define a new variable based on a selection
poker_midweek <- poker_vector[index_midweek]
```

*** =sct
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_midweek", 
            incorrect_msg = "It looks like `poker_midweek` does not contain the correct values from `poker_vector`. You can use the vector `c(2, 3, 4)` inside square brackets.")
success_msg("Well done! Continue to the next exercise to specialize in vector selection some more!");
```


--- type:NormalExercise xp:100 skills:1 key:27976b79f4
## Vector selection: the good times (3)

Sometimes rather than specifying the elements that you want to select, it is easier to specify the elements that you **don't want** to select. In this case, rather than writing a vector of positive numbers inside the square brackets, you write a vector of **negative numbers**.

For example, since `poker_vector` has five elements, both these lines of code will drop the first two elements.

```{r}
poker_vector[-1:-2]
## Wednesday  Thursday    Friday 
##     20.25   -120.00    240.00
poker_vector[3:5]
## Wednesday  Thursday    Friday 
##     20.25   -120.00    240.00
```

*** =instructions

Thursday was a bad day for roulette. Ignore Thursday by using negative indexing to drop the fourth day's result.

*** =hint
Assign a selection of `roulette_vector` to `roulette_selection_vector` by placing `2:5` between square brackets.

*** =sample_code
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Define a new variable based on a selection
roulette_ignoring_thursday <- 
```

*** =solution
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Define a new variable based on a selection
roulette_ignoring_thursday <- roulette_vector[-4]
```

*** =sct

```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_ignoring_thursday", 
            undefined_msg = "Please make sure to define a variable `roulette_selection_vector`.",
            incorrect_msg = "It looks like `roulette_ignoring_thursday` does not contain the correct selection from `roulette_vector`. Make sure to to use the right indexes.")
success_msg("Awesome! The colon operator is extremely useful and very often used in R programming, so remember it well. Proceed to the next exercise.")
```


--- type:NormalExercise xp:100 skills:1 key:e6c263ddee
## Vector selection: the good times (4)

If you have a named vector, you can also select elements of a vector using its names. This often makes your code easier to read (at the cost of a little more typing).

For example, both these piece of code select the poker elements for Tuesday and Thursday, but using names makes the code easier to understand. (The output is shown in comments.)

```{r}
poker_vector[c(2, 4)]
##  Tuesday Thursday 
##    -50.5   -120.0
poker_vector[c("Tuesday", "Thursday")]
##  Tuesday Thursday 
##    -50.5   -120.0
```

The (arithmetic) **mean** is a very common type of average. It is calculated by adding up all the numbers in the vector, then dividing by the length of that vector. You can calculate it using the [`mean()`](https://www.rdocumentation.org/packages/base/topics/mean) function.  For example

```{r}
mean(c(4, 6, 7, 3, 5))
## [1] 5
```

*** =instructions
- Select the first three elements in `roulette_vector` by using their names: `"Monday"`, `"Tuesday"` and `"Wednesday"`. Assign the result of the selection to `roulette_start`.
- Calculate the average of the values in `roulette_start` with the [`mean()`](http://www.rdocumentation.org/packages/base/functions/mean) function. Simply print out the result so you can inspect it.

*** =hint
- You can use `c("Monday", "Tuesday", "Wednesday")` inside square brackets to subset `roulette_vector` appropriately.
- You can use `mean(roulette_start)` to get the mean of the elements in `roulette_start`. You do not need the mean of all roulette elements, but only of the first three days.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Select roulette results for Monday, Tuesday and Wednesday
roulette_start <- 
  
# Calculate the average of the elements in roulette_start

```

*** =solution
```{r}
# Poker and roulette winnings from Monday to Friday:
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Select roulette results for Monday, Tuesday and Wednesday
roulette_start <- poker_vector[c("Monday", "Tuesday", "Wednesday")]
  
# Calculate the average of the elements in roulette_start
mean(roulette_start)
```

*** =sct
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_start", 
            incorrect_msg = "It looks like `poker_start` does not contain the first three values of `roulette_vector`. You can use `c(\"Monday\", \"Tuesday\", \"Wednesday\")` inside square brackets to do this.")
test_output_contains("mean(roulette_start)", incorrect_msg = "Have you correctly calculated the average of the values in `roulette_start` and printed it out? Use `mean(roulette_start)`.")
success_msg("Good job! Apart from subsetting vectors by index or by name, you can also subset vectors by comparison. The next exercises will show you how!");
```



--- type:NormalExercise xp:100 skills:1 key:2754fc5cd4
## Selection by comparison

So far, you've used positive integers, negative integers, and names to select elements from a vector. The fourth (and final) way of accessing elements from a vector is to use **logical** values.

A common scenario is to use the comparison operators that you saw in the Calculating Total Winnings exercise to select values.  For example, you can ignore all your losses and ask R to return only those days where you realized a positive return for poker (the output is shown in comments).

```{r}
poker_vector > 0
##    Monday   Tuesday Wednesday  Thursday    Friday 
##      TRUE     FALSE      TRUE     FALSE      TRUE
poker_vector[poker_vector > 0]
##    Monday Wednesday    Friday 
##    140.25     20.25    240.00
```

As you may have spotted in the example, when you use a logcial vector for selection, R selects the elements that are `TRUE`.

*** =instructions

* Use a comparison operator to create a logical vector that is `TRUE` when `roulette_vector` is greater than zero, and assign the result to `selection_vector`.
* Use `selection_vector` in square brackets to assign the amounts that you won on the profitable days to the variable `roulette_winning_days`.

*** =hint

* Use `roulette_vector > 0` to create a logical vector that is `TRUE` when `roulette_vector` is greater than zero.
* Use `roulette_vector[selection_vector]` to select the desired elements from `roulette_vector`, and assign the result to `roulette_winning_days`.

*** =pre_exercise_code
```{r}
# no pec
```

*** =sample_code
```{r}
# Poker and roulette winnings from Monday to Friday
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Which days did you make money on roulette?
selection_vector <- 

# Select these days from roulette_vector
roulette_winning_days <- 
```

*** =solution
```{r}
# Poker and roulette winnings from Monday to Friday
poker_vector <- c(Monday = 140.25, Tuesday = -50.50, Wednesday = 20.25, Thursday = -120, Friday = 240)
roulette_vector <- c(Monday = -24.75, Tuesday = -50.25, Wednesday = 100, Thursday = -350.50, Friday = 10)

# Which days did you make money on roulette?
selection_vector <- roulette_vector > 0

# Select these days from roulette_vector
roulette_winning_days <- roulette_vector[selection_vector]
```

*** =sct
```{r}
msg = "Do not change anything about the definition and naming of `poker_vector` and `roulette_vector`."
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("selection_vector", incorrect_msg = "Don't change the way `selection_vector` is calculated.")
test_object("poker_winning_days",
            incorrect_msg =  "It looks like `poker_winning_days` does not contain the correct result. Use `poker_vector[selection_vector]`.")
success_msg("Good job! Continue to the next exercise.")
```


