--- 
title_meta  : Chapter 1
title       : Intro to basics
description : In this chapter you will take your first steps with R. You learn how to use the console as a calculator, assign variables, and get to know the basic data types in R. Let's get started!
framework   : datacamp
mode        : selfcontained
---


## How it works

In the **editor** on the right you should type R code to solve the exercises. 

R makes use of the `#` sign to add comments, such that you and others can understand what the R code is about. Just like Twitter! 
Comments are not run as R-code, so they will not influence your result. 

The output of your R code is shown in the **console** in the lower right corner, while **graphs** are shown in the upper right corner.

*** =instructions
- Click Submit Answer and see how the console now shows you the executed R code: the solution `7` appears as the sum of `3` and `4`. In its most basic form R can thus be used as a calculator or to generate plots, but there's much more ;-).
- In the upper right corner R generated some cool visualizations. Use the arrows above the graphs to browse through these. 

*** =hint
Just click the Submit Answer button.

*** =sample_code

```{r eval=FALSE}
# This  is the editor and the part below the white line is called the console

# The  hashtag is used to add comments

# Show some demo graphs generated with R
demo("graphics")

# Calculate 3+4
3+4
```

*** =solution

```{r eval=FALSE}
#Just click the Submit Answer button

# This  is the editor and the part below the white line is called the console

# The  hashtag is used to add comments

# Calculate 3+4
3+4

# Show some demo graphs generated with R
demo("graphics")
```

*** =sct

```{r eval=FALSE}
DM.result = list(TRUE, "Great! Let's get started!");
```

*** =pre_exercise_code

```{r eval=FALSE}

```


---

## Little arithmetics with R

In its most basic form R can be used as a simple calculator. Let's consider the following arithmetic operators:

- Addition: `+`
- Subtraction: `-`
- Multiplication: `*`
- Division: `/`
- Exponentiation: `^`
- Modulo: `%%`

The last two might need some explaining:
- The `^` operator takes the number left to it to the power of the number on its right-hand side: for example `3^2` is 9.
- The modulo returns the remainder of the division of the number left to it by the number on its right-hand side, for example 5 modulo 3 or `5%%3` is 2.

*** =instructions
- Type `2^5` in the editor to calculate 2 to the power 5.
- Type `17%%4` to calculate 17 modulo 4.
- Click Submit Answer and have a look at the R output in the console.
- Note how the `#` symbol is used to give comments on certain lines.

*** =hint
Just follow the instructions, this should be easy ;-).

*** =sample_code

```{r eval=FALSE}
# An addition
5+5 

# A subtraction
5-5 

# A multiplication
3*5

 # A division
(5+5)/2 

# Exponentiation

# Add your code here!

# Modulo

# Add your code here!
```

*** =solution

```{r eval=FALSE}
# An addition
5+5 

# A subtraction
5-5 

# A multiplication
3*5

 # A division
(5+5)/2 

# Exponentiation
2^5
# Modulo
17%%4
```

*** =sct

```{r eval=FALSE}
if (!(isTRUE(try(output_contains("2^5"))))) {
  DM.result = list(FALSE, "The exponentiation example doesn't seem right. Make sure to carefully read the instructions.");
} else if (!(isTRUE(try(output_contains("17%%4"))))) {
  DM.result = list(FALSE, "There seems to be an issue with the modulo example. Make sure to carefully read the instructions.");
} else {
  DM.result = list(TRUE, "Great! Continue to the next exercise.");
}
```

*** =pre_exercise_code

```{r eval=FALSE}

```

---

## Variable assignment 

A basic concept in (statistical) programming is called a **variable**. 

A variable allows you to store a value (e.g. 4) or object (e.g. a function description) in R. Then later you can use this variable's name to easily access the value or object that is stored within this variable. 

You can assign a value 4 to a variable `my_variable` with the command: `my_variable = 4`. 

*** =instructions
- Up to you: Complete the code in the editor such that it assigns the value 42 to the variable `x` in the editor. Click Submit Answer. Notice that when asking R to print `x` the value 42 appears.

*** =hint

Look at how the value 4 was assigned to `my_variable`.

*** =sample_code

```{r eval=FALSE}
# Assign the value 42 to x
x = 

# Print out the value of the variable x
x
```

*** =solution

```{r eval=FALSE}
# Assign the value 42 to x
x = 42

# Print out the value of the variable x
x
```

*** =sct

```{r eval=FALSE}
if (!exists("x")){
  DM.result = list(FALSE, "Please make sure to define a variable 'x'.");
} else if (!isTRUE(try(all.equal(x,42)))){
  DM.result = list(FALSE, "Make sure that you assign the correct value to 'x'.");
} else {
	DM.result = list(TRUE, "Good job! Continue to the next exercise.");
}
```

*** =pre_exercise_code
```{r eval=FALSE}

```


---

## Variable assignment (2)

Suppose you have a fruit basket with 5 apples. As a data analyst in training, you want to store the number of apples in a variable with the name `my_apples`. 

*** =instructions
- Type the following code: `my_apples = 5` to assign the value 5 to `my_apples`.
- Type: `my_apples` below the second comment.
- Click Submit Answer, and look at the console: you see that the number 5 is printed. So R now links the variable `my_apples` to the value 5.

*** =hint

Remember, if you want to assign a number or object to a variable in R, you can make use of the assignment operator `=`.

*** =sample_code

```{r eval=FALSE}
# Assign the value 5 to the variable called my_apples

# Print out the value of the variable my_apples
```

*** =solution

```{r eval=FALSE}
# Assign the value 5 to the variable called my_apples
my_apples = 5

# Print out the value of the variable my_apples
my_apples
```

*** =sct

```{r eval=FALSE}
if (!exists("my_apples")){
  DM.result = list(FALSE, "Please make sure to define a variable 'my_apples'.");
} else if (!isTRUE(try(all.equal(my_apples,5)))){
  DM.result = list(FALSE, "Make sure that you assign the correct value to 'my_apples'.");
} else {
	DM.result = list(TRUE, "Good job! Continue to the next exercise.");
}
```

*** =pre_exercise_code
```{r eval=FALSE}

```

---

## Variable assignment (3)

Every tasty fruit basket needs oranges so you decide to add 6 oranges. As a data analyst your reflex is to immediately create the variable `my_oranges` and assign the value 6 to it. Next, you want to calculate how many pieces of fruit you have in total. Since you have given meaningful names to these values, you can now code this in a clear way: `my_apples + my_oranges`.

*** =instructions
- Assign to `my_oranges` the value 6.
- R allows you to combine the variables `my_apples` and `my_oranges` into a new variable `my_fruit`. Create this new variable `my_fruit`, which is the total amount of fruits in your fruit basket.

*** =hint
`my_fruit` is just the sum of `my_apples` and `my_oranges`. You can use the `+` operator to sum the two and `=` to assign that value to the variable `my_fruit`.

*** =sample_code

```{r eval=FALSE}
# Assign a value to the variables called my_apples and my_oranges
my_apples = 5

# Add these two variables together and print the result

# Create the variable my_fruit
```

*** =solution

```{r eval=FALSE}
# Assign a value to the variable called my_apples and my_oranges
my_apples  = 5
my_oranges = 6

# Add these two variables together and print the result
my_apples + my_oranges

# Create the variable my_fruit by adding my_apples and my_oranges
my_fruit = my_apples + my_oranges
```

*** =sct

```{r eval=FALSE}
if (!exists("my_oranges")) {
  DM.result = list(FALSE, "Please make sure to define a variable 'my_oranges'.");
} else if (!isTRUE(try(all.equal(my_oranges,6)))) {
  DM.result = list(FALSE, "Make sure that you assign the correct value to 'my_oranges'.");
  DM.result = list(FALSE, "The output doesn't contain the result of adding 'my_apples' and 'my_oranges'.");
} else if (!exists("my_fruit")) {
  DM.result = list(FALSE, "Please make sure to define a variable 'my_fruit'.");
} else if (!isTRUE(try(all.equal(my_fruit,my_apples + my_oranges)))) {
  DM.result = list(FALSE, "Make sure that you assign the correct value to 'my_fruit'.");
} else {
  DM.result = list(TRUE, "Good job! Continue to the next exercise.");
}
```

*** =pre_exercise_code

```{r eval=FALSE}

```

---

## Apples and oranges

Common knowledge tells you not to add apples with oranges. But hey, that's what you just did, didn't you :-)? The `my_apples` and `my_oranges` variables both contained a number in the previous exercise. The `+` operator works with what's called numeric variables in R. Suppose you really tried to add "apples" and "oranges", and assigned a text value to the variable `my_oranges` (see the editor). You now try to assign the addition of a numeric and a character variable to the variable `my_fruit`, which is not possible.

*** =instructions
- Click Submit Answer and read the error message. Make sure to understand why this didn't work.
- Adjust the code such that R knows you have 6 oranges and thus a fruit basket with 11 pieces of fruit.

*** =hint
You have to assign the numeric value `6` to the `my_oranges` variable instead of the character value `"six"`. Note how the quotation marks are used to indicate that `"six"` is a character.

*** =sample_code

```{r eval=FALSE}
# Assign a value to the variable called my_apples
my_apples = 5 
# Print out the value of my_apples
my_apples       

# Add a character
my_oranges = "six" 
my_oranges 

# New variable that contains total amount of fruit
my_fruit   = my_apples + my_oranges 
my_fruit
```

*** =solution

```{r eval=FALSE}
# Assign a value to the variable called my_apples
my_apples = 5  

# Print out the value of answer
my_apples  

# Add a numeric
my_oranges = 6
my_oranges 

# New variable that contains total amount of fruit
my_fruit   = my_apples + my_oranges 
my_fruit
```

*** =sct

```{r eval=FALSE}
if (!exists("my_oranges")) {
  DM.result = list(FALSE, "Please make sure to define a variable 'my_oranges'.");
} else if (!isTRUE(try(all.equal(my_oranges,6)))) {
  DM.result = list(FALSE, "Make sure that you assign the correct value to 'my_oranges'. A string does not equal a number!");
} else if (!(isTRUE(try(output_contains("my_apples + my_oranges"))))) {
  DM.result = list(FALSE, "The output doesn't contain the result of adding 'my_apples' and 'my_oranges'.");
} else if (!exists("my_fruit")) {
  DM.result = list(FALSE, "Please make sure to define a variable 'my_fruit'.");
} else if (!isTRUE(try(all.equal(my_fruit,my_apples + my_oranges)))) {
  DM.result = list(FALSE, "Make sure that you assign the correct value to 'my_fruit'.");
} else {
  DM.result = list(TRUE, "Good job! Continue to the next exercise.");
}
```

*** =pre_exercise_code

```{r eval=FALSE}

```


---

## Basic data types in R

R works with numerous data types. Some of the most basic types to get started are:

- Decimals values like `4.5` are called **numerics**.
- Natural numbers like `4` are called **integers**.
- Boolean values (`TRUE` or `FALSE`) are called **logical** (`TRUE` can be abbreviated to `T` and `FALSE` to `F`).
- Text (or string) values are called **characters**.

Note how the quotation marks on the right indicate that "some text" is a character.

*** =instructions
Change the value of:
- `my_numeric` variable to `42`.
- `my_character` variable to `"forty-two"`. Note: the quotation marks indicate that `"forty-two"` is a character.
- `my_logical` variable to `FALSE`.

Note that R is case sensitive!

*** =hint 
Replace in the editor the values with the values provided in the exercise. For example: 
`my_numeric = 42` assigns the value 42 to the variable `my_numeric`. 

*** =sample_code

```{r eval=FALSE}
# What's the answer to the universe
my_numeric = 42.5

# The quotation marks indicate that the variable is of type character
my_character = "some text"

my_logical = TRUE
```

*** =solution

```{r eval=FALSE}
# What's the answer to the universe?
my_numeric = 42

# The quotation marks indicate that the variable is of type character
my_character = "forty-two"

my_logical = FALSE
```

*** =sct

```{r eval=FALSE}
if (!exists("my_numeric")) {
  DM.result = list(FALSE, "Please make sure to define a variable 'my_numeric'.");
} else if (!isTRUE(try(all.equal(my_numeric,42)))) {
  DM.result = list(FALSE, "Make sure that you assign the correct value to 'my_numeric.'");
} else if (!exists("my_character")) {
  DM.result = list(FALSE, "Please make sure to define a variable 'my_character'.");
} else if (!isTRUE(try(all.equal(my_character,"forty-two")))) {
  DM.result = list(FALSE, "Make sure that you assign the correct value to 'my_character'. Don't forget the quotes.");
} else if (!exists("my_logical")) {
  DM.result = list(FALSE, "Please make sure to define a variable 'my_logical'.");
} else if (!isTRUE(try(all.equal(my_logical,FALSE)))) {
  DM.result = list(FALSE, "Make sure that you assign the correct value to 'my_logical'.");
} else {
  DM.result = list(TRUE, "Good job! Continue to the next exercise.");
}
```

*** =pre_exercise_code
```{r eval=FALSE}

```


---

## What's that data type?

Remember when you added `5 + "six"` and got an error due to a mismatch in data types? You avoid such embarrassing situations by checking the data type of a variable upfront. You can do this as follow: `class(some_variable_name)`.

*** =instructions
- Complete the code in the editor and print the class of `my_numeric`, `my_character` and `my_logical` to the console. 

*** =hint
You can find out the data type of the `my_numeric` variable for example by typing: `class(my_numeric)`. What you should do is similar for `my_character` and `my_logical`. 

*** =sample_code

```{r eval=FALSE}
# Declare variables of different types
my_numeric = 42
my_character = "forty-two"
my_logical = FALSE 

# Check which type these variables have:

```

*** =solution

```{r eval=FALSE}
# Declare variables of different types:
my_numeric = 42
my_character = "forty-two"
my_logical = FALSE

# Check which type these variables have:
class(my_numeric)
class(my_character)
class(my_logical)
```

*** =sct

```{r eval=FALSE}
if (!(isTRUE(try(output_contains("class(my_numeric)"))))) {
  DM.result = list(FALSE, "The output doesn't contain the class of 'my_numeric'.");
} else if (!(isTRUE(try(output_contains("class(my_character)"))))) {
  DM.result = list(FALSE, "The output doesn't contain the class of 'my_character'.");
} else if (!(isTRUE(try(output_contains("class(my_logical)"))))) {
  DM.result = list(FALSE, "The output doesn't contain the class of 'my_logical'.");
} else {
	DM.result = list(TRUE, "Congratulations! This was the last exercise for this chapter. Click next to go back to the overview.");
}
```

*** =pre_exercise_code
```{r eval=FALSE}

```
