--- 
title_meta  : Chapter 5
title       : Data frames
description : Most data sets you will be working with will be stored as a data frame. By the end of this chapter, you will be able to create a data frame, select interesting parts of a data frame and order a data frame according to certain variables. 
framework   : datacamp
mode        : selfcontained

---
## What's a data frame?

You may remember from the chapter about matrices that all the elements that you put in a matrix should be of the same type. Back then, your data set on Star Wars only contained numeric elements. 

When doing a market research survey, however, you often have questions such as:
- 'Are your married?' or 'yes/no' questions (= boolean data type)
- 'How old are you?' (= numeric data type)
- 'What is your opinion on this product?' or other 'open-ended' questions (= character data type)
- ...

The output, namely the respondents' answers to the questions formulated above, is a data set of different data types. You will often find yourself working with data sets that contain different data types instead of only one. 

A data frame has the variables of a data set as columns and the observations as rows. This will be a familiar concept for those coming from different statistical software packages such as SAS or SPSS.

*** =instructions
Click 'Submit Answer'. The data from the built-in example data frame [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars) will be printed to the console.

*** =hint
Just click 'Submit Answer' and witness the magic!

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
mtcars # Built-in R data set stored in a data frame
```

*** =solution
```{r eval=FALSE}
mtcars # Built-in R data set stored in a data frame<
```

*** =sct
```{r eval=FALSE}
test_output_contains("mtcars", incorrect_msg = "Do not change anything about the code, Make sure that you output <code>mtcars</code>.");
success_msg("Great! Continue to the next exercise.")
```

---
## Quick, have a look at your data set

Wow, that is a lot of cars! 

Working with large data sets is not uncommon in data analysis. When you work with (extremely) large data sets and data frames, your first task as a data analyst is to develop a clear understanding of its structure and main elements. Therefore, it is often useful to show only a small part of the entire data set. 

So how to do this in R?
Well, the function [`head()`](http://www.rdocumentation.org/packages/utils/functions/head) enables you to show the first observations of a data frame (or any R object you pass to it). Unoriginally, the function [`tail()`](http://www.rdocumentation.org/packages/utils/functions/head) prints out the last observations in your data set.

Both [`head()`](http://www.rdocumentation.org/packages/utils/functions/head) and [`tail()`](http://www.rdocumentation.org/packages/utils/functions/head) print a top line called the 'header', which contains the names of the different variables in your data set.

*** =instructions
Give an overview of the [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars) data set by printing its first observations and the corresponding header.

*** =hint
Use the [`head()`](http://www.rdocumentation.org/packages/utils/functions/head) function and ask the head of [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars).

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Have a quick look at your data

```

*** =solution
```{r eval=FALSE}
# Have a quick look at your data
head(mtcars)
```

*** =sct
```{r eval=FALSE}
test_function("head", "x", incorrect_msg = "Have you specified the correct parameter in the <code>head()</code> function? Make sure to pass it a data set you want to inspect, <code>mtcars</code> in this case.")
test_output_contains("head(mtcars)", 
                     incorrect_msg = "Make sure that you use the <code>head()</code> function on <code>mtcars</code>.")
success_msg("Wonderful! So, what do we have in this data set? For example, <code>hp</code> for example represents the car's horsepower; the Datsun has the lowest horse power of the 6 cars that are displayed. For a full overview of the variables' meaning, type <code>?mtcars</code> in the console and read the help page. Continue to the next exercise!");
```

---
## Have a look at the structure

Another method that is often used to get a rapid overview of your data is the function [`str()`](http://www.rdocumentation.org/packages/utils/functions/str). The function [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) shows you the structure of your data set. For a data frame it tells you:

- The total number of observations (e.g. 32 car types)
- The total number of variables (e.g. 11 car features)
- A full list of the variables names (e.g. mpg, cyl ... )
- The data type of each variable (e.g. num for car features)
- The first observations

Applying the [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) function will often be the first thing that you do when receiving a new data set or data frame. It is a great way to get more insight in your data set before diving into the real analysis.

*** =instructions
Investigate the structure of [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars). Make sure that you see the same numbers, variables and data types as mentioned above.

*** =hint
Use the [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) function with [`mtcars`](http://www.rdocumentation.org/packages/datasets/functions/mtcars) as input!

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Investigate the structure of the mtcars data set to get started!
```

*** =solution
```{r eval=FALSE}
# Investigate the structure of the mtcars data set to get started!
str(mtcars)
```

*** =sct
```{r eval=FALSE}
test_function("str","object",incorrect_msg = "Make sure to check the structure of the <code>mtcars</code> data set.")
test_output_contains("str(mtcars)", incorrect_msg = "Make sure that you use the <code>str()</code> function on <code>mtcars</code>.")
success_msg("Nice work! Can you find all the information that is listed in the exercise's assignment? Continue to the next exercise.")
```

---
## Creating a data frame 

Since using built-in data sets is not even half the fun of creating your own data sets, the rest of this chapter is based on your personally developed data set. So put your jet pack on because it is time for some good old fashioned space exploration! 

As a first goal, you want to construct a data frame that describes the main characteristics of eight planets in our solar system. According to your good friend Buzz, the main features of a planet are:

- The type of planet (Terrestrial or Gas Giant).
- The planet's diameter relative to the diameter of the Earth.
- The planet's rotation across the sun relative to that of the Earth.
- If the planet has rings or not (TRUE or FALSE).

After doing some high-quality research on [Wikipedia](http://en.wikipedia.org/wiki/Planet), you feel confident enough to create the necessary vectors: `planets`, `type`, `diameter`, `rotation` and `rings` (see editor, note that an element of one vector is linked to the element of another vector based on its position).

You construct a data frame with the [`data.frame()`](http://www.rdocumentation.org/packages/base/functions/data.frame) function. As arguments, you should provide the above mentioned vectors as input that should become the different columns of that data frame. Therefore, it is important that each vector used to construct a data frame has an equal length. But do not forget that it is possible (and likely) that they contain different types of data.

*** =instructions
Use the function [`data.frame()`](http://www.rdocumentation.org/packages/base/functions/data.frame) to construct a data frame. Call this variable `planets_df`.

*** =hint
The `data.frame(col1, col2, col3,...)` function takes as arguments the vectors that will become the columns of the data frame. The columns in this case are (in this order): `planet`, `type`, `diameter`, `rotation` and `rings`.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune");
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883); 
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67);
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE);

# Create the data frame:
planets_df  <-

```

*** =solution
```{r eval=FALSE}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune");
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883); 
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67);
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE);

# Create the data frame:
planets_df 	<- data.frame(planets,type,diameter,rotation,rings)

```

*** =sct
```{r eval=FALSE}

test_correct({
  test_object("planets_df",
              undefined_msg = "Please make sure to define a variable <code>planets_df</code>.",
              incorrect_msg = "Make sure to assign the correct order of arguments to the data.frame <code>planets_df</code>. The correct order is planets, type, diameter, rotation and rings.")  
}, {
  msg = "Do not change anything about the definition of the vector. Only add code to create the <code>planets_df</code> data frame!"
  test_object("planet", undefined_msg = msg, incorrect_msg = msg)
  test_object("type", undefined_msg = msg, incorrect_msg = msg)
  test_object("diameter", undefined_msg = msg, incorrect_msg = msg)
  test_object("rotation", undefined_msg = msg, incorrect_msg = msg)
  test_object("rings", undefined_msg = msg, incorrect_msg = msg)
})

success_msg("Great job! Continue to the next exercise. The logical next step, as you know by now, is inspecting the data frame you just created. Head over to the next exercise.");
```

---
## Creating a data frame (2)

Make sure that you have 8 observations and 5 variables. The `planets_df` data frame has been made available in the workspace, so you can directly use it.

*** =instructions
Use the function [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) to investigate the structure of the new `planets_df` variable.

*** =hint
This one is easy, no hints this time!

*** =pre_exercise_code
```{r eval=FALSE}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune");
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883); 
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67);
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE);
planets_df <- data.frame(planets,type,diameter,rotation,rings)
```

*** =sample_code
```{r eval=FALSE}
# Check the structure of 'planets_df'
```

*** =solution
```{r eval=FALSE}
# Check the structure of 'planets.df'
str(planets_df)
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or overwrite the <code>planets_df</code> data frame!"
test_object("planets_df", undefined_msg = msg, incorrect_msg = msg)
test_function("str", "object", incorrect_msg = "Make sure that you use the <code>str()</code> function on <code>planets_df</code>.")
success_msg("Awesome! Now that you have a clear understanding of the <code>planets_df</code> data set, it's time to see how you can select elements from it. Learn all about in the next exercises!")
```

---
## Selection of data frame elements

Similar to vectors and matrices, you select elements from a data frame with the help of square brackets `[ ]`. By using a comma, you can indicate what to select from the rows and the columns respectively. For example:

- `my_data_frame[1,2]` selects from the first row in `my_data_frame` the second element.
- `my_data_frame[1:3,2:4]` selects rows 1,2,3 and columns 2,3,4 in `my_data_frame`.

Sometimes you want to select all elements of a row or column. To do this, you can use the follwing. `my_data_frame[1, ]` selects all elements of the first row. Let us now apply this technique on `planets_df`!

*** =instructions
- Create the data frame `closest_planets_df` that contains all data on the first three planets.
- Create the data frame `furthest_planets_df` that contains all data on the last three planets.

*** =hint
`planets_df[1:3,]` will select alle elements of the first three rows.

*** =pre_exercise_code
```{r eval=FALSE}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune");
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883); 
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67);
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE);
planets_df   <- data.frame(planets,type,diameter,rotation,rings)
```

*** =sample_code
```{r eval=FALSE}
# The 'planets_df' data frame from the previous exercise is pre-loaded

# All data from the first three planets
closest_planets_df <- 

# All data from the last three planets
furthest_planets_df <- 

# Have a look:
closest_planets_df
furthest_planets_df
```

*** =solution
```{r eval=FALSE}
# The 'planets_df' data frame from the previous exercise is pre-loaded

# All data from the first three planets
closest_planets_df <- planets_df[1:3, ]

# All data from the last three planets
furthest_planets_df <- planets_df[6:8, ]

# Have a look:
closest_planets_df
furthest_planets_df
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or overwrite the <code>planets_df</code> data frame!"
test_object("planets_df", undefined_msg = msg, incorrect_msg = msg)
test_object("closest_planets_df",
            undefined_msg = "Please make sure to define a variable <code>closest_planets_df</code>.",
            incorrect_msg = "Did you select the first three rows of <code>planets_df</code>?")
test_object("furthest_planets_df",
            undefined_msg = "Please make sure to define a variable <code>furthest_planets_df</code>.",
            incorrect_msg = "Make sure that you select the last three rows of <code>planets_df</code>.")
success_msg("Great! Apart from selecting elements from your data frame by index, you can also use the column names. To learn how, head over to the next exercise.")
```

---
## Selection of data frame elements (2)

Instead of using numerics to select elements of a data frame, you can also use the variable names to select columns of a data frame. 

Maybe you want to select only the first 3 elements of the variable 'type'. One way to do this is: `planets_df[1:3,1]`. A possible disadvantage of this approach is that you have to know (or look up) the position of the variable 'type', which gets hard if you have a lot of variables. It is often easier to just make use of the variable name, e.g. `planets_df[1:3,"type"]`.

*** =instructions
Select for the last six rows only the diameter and assign this selection to `furthest_planets_diameter`.

*** =hint
You select elements of a data frame conveniently with the square brackets. Select `3:8` for the rows, and `"diameter"` for the columns.

*** =pre_exercise_code
```{r eval=FALSE}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune");
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883); 
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67);
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE);
planets_df   <- data.frame(planets,type,diameter,rotation,rings)
```

*** =sample_code
```{r eval=FALSE}
# The 'planets_df' data frame from the previous exercise is pre-loaded

# Do selection on both rows and columns
furthest_planets_diameter <- 
```

*** =solution
```{r eval=FALSE}
# The planets_df data frame from the previous exercise is pre-loaded

# Do selection on both rows and columns
furthest_planets_diameter <- planets_df[3:8,"diameter"]
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or overwrite the <code>planets_df</code> data frame!"
test_object("planets_df", undefined_msg = msg, incorrect_msg = msg)
test_object("furthest_planets_diameter",
            undefined_msg = "Please make sure to define a variable <code>furthest_planets_diameter</code>.",
            incorrect_msg = "Make sure that you select the last six rows and only the <code>diameter</code> column.")
success_msg("Nice! Continue to the next exercise!")
```

---
## Only planets with rings

You will often want to select an entire column, namely one specific variable from a data frame. If you want to select all elements of the variable `rings`, both `planets_df[,5]` and `planets_df[,"rings"]` do the trick.

However, there is a short-cut. Use the `$`-sign to tell R that it only has to look up all the elements of the variable behind the sign: 

```
data_frame_name$variable_name
```

*** =instructions
- Make use of the `$`-sign to create the variable `rings_vector` that contains all elements of the `rings` variable in the `planets_df` data frame.

*** =hint
`data_frame_name$variable_name` is the most convenient way to select a variable from a data frame. In this case, the data frame of choice is `planets_df` and the variable of choice is `rings`.

*** =pre_exercise_code
```{r eval=FALSE}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune");
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883); 
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67);
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE);
planets_df   <- data.frame(planets,type,diameter,rotation,rings)
```

*** =sample_code
```{r eval=FALSE}
# 'planets_df' is pre-loaded in your workspace

# Create the rings_vector
rings_vector <- 
```

*** =solution
```{r eval=FALSE}
# 'planets_df' is pre-loaded in your workspace

# Create the rings_vector
rings_vector <- planets_df$rings
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or overwrite the <code>planets_df</code> data frame!"
test_object("planets_df", undefined_msg = msg, incorrect_msg = msg)
test_object("rings_vector",
            undefined_msg = "Please make sure to define a variable <code>rings_vector</code>.",
            incorrect_msg = "It looks like <code>rings_vector</code> does not contain all the element of the <code>ring</code> variable of the <code>planets_df</code>.")
success_msg("Great! Continue to the next exercise and discover yet another way of subsetting!")
```

---
## Only planets with rings (2)

You probably remember from highschool that some planets in our solar system have rings and others do not. But due to other priorities at that time (read: puberty) you can not recall their names, let alone their rotation speed, etc. 

Could R help you out? (Spoiler alert: of course it can!) 

If you type `rings_vector` in the console, you get:

```
[1] FALSE FALSE FALSE FALSE  TRUE  TRUE  TRUE  TRUE
```

This means that the first four observations (or planets) do not have a ring (`FALSE`), but the other four do (`TRUE`). However, you do not get a nice overview of the names of these planets, their diameter, etc. As a next step, use `rings_vector` to select all the data (that is, all columns) on the four planets with rings from `planets_df`.

*** =instructions
Assign to `planets_with_rings_df` all data in the `planets_df`data set for the planets with rings, that is, where `rings_vector` is `TRUE`.

*** =hint
Select elements from `planets_df` by using the square brackets. The `rings_vector` contains boolean values and R will select only those rows/columns were the vector element is `TRUE`. In this case, you want to select rows based on `rings_vector` and select all the columns.

*** =pre_exercise_code
```{r eval=FALSE}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune");
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883); 
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67);
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE);
planets_df   <- data.frame(planets,type,diameter,rotation,rings)
rings_vector <- planets_df$rings
```

*** =sample_code
```{r eval=FALSE}
# 'planets_df' and 'rings_vector' are pre-loaded in your workspace

# Select the information on planets with rings:
planets_with_rings_df <- 
```

*** =solution
```{r eval=FALSE}
# 'planets_df' and 'rings_vector' are pre-loaded in your workspace

# Select the information on planets with rings:
planets_with_rings_df <- planets_df[rings_vector,]
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or overwrite <code>planets_df</code> or <code>rings_vector</code>!"
test_object("planets_df", undefined_msg = msg, incorrect_msg = msg)
test_object("rings_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("planets_with_rings_df",
            undefined_msg = "Please make sure to define a variable <code>planets_with_rings_df</code>.",
            incorrect_msg = "It looks like <code>planets_with_rings_df</code> does not contain all the data of the planets with rings.");
success_msg("Wonderful! This is a rather tedious solution. The next exercise will teach you how to do it in a more concise way.")
```

---
## Only planets with rings but shorter

So what exactly did you learn in the previous exercises? You selected a subset from a data frame (`planets_df`) based on whether or not a certain condition was true (rings or no rings), and you managed to pull out all relevant data. Pretty awesome! By now, NASA is probably already flirting with your CV ;-). 

Now, let us move up one level and use the function [`subset()`](http://www.rdocumentation.org/packages/base/functions/subset). You should see the [`subset()`](http://www.rdocumentation.org/packages/base/functions/subset) function as a short-cut to do exactly the same as what you did in the previous exercises. 

```
subset(my_data_frame, subset = some_condition)
``` 

The first argument of [`subset()`](http://www.rdocumentation.org/packages/base/functions/subset) specifies the data set for which you want a subset. By adding the second argument, you give R the necessary information and conditions to select the correct subset. 

For example: 
```
subset(planets_df, subset = planets_df$rings == TRUE)
``` 

R will give you exactly the same result as you got in the previous exercise. But this time without needing the `rings_vector`!

*** =instructions
Create a dataframe `small_planets_df` with planets that have a diameter smaller than the Earth (so smaller than 1, since diameter is a relative measure of the planet's diameter w.r.t that of planet Earth).

*** =hint
Sorry, you are on your own here.

*** =pre_exercise_code
```{r eval=FALSE}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune");
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883); 
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67);
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE);
planets_df   <- data.frame(planets,type,diameter,rotation,rings)
```

*** =sample_code
```{r eval=FALSE}
# 'planets_df' is pre-loaded in your workspace

# Planets that are smaller than planet Earth:
small_planets_df  <- # Your code here!
```

*** =solution
```{r eval=FALSE}
# 'planets_df' is pre-loaded in your workspace

# Planets that are smaller than planet Earth:
small_planets_df  <- subset(planets_df, subset = planets_df$diameter < 1)
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or overwrite the <code>planets_df</code> data frame!"
test_object("planets_df", undefined_msg = msg, incorrect_msg = msg)
test_object("small_planets_df",
            undefined_msg = "Please make sure to define a variable <code>small_planets_df</code>.",
            incorrect_msg = "It looks like <code>small_planets_df</code> does not contain all the correct subset of planets_df.");
success_msg("Great! Not only is the <code>subset()</code> function more concise, it is probably also more understandable for people who read your code. Continue to the next exercise.");
```

---
## Sorting

Making and creating rankings is one of mankind's favorite affairs. These rankings can be useful (best universities in the world), entertaining (most influencial moviestars) or pointless (best 007 look-a-like). Up to you for which purpose you want to use your R skills ;-). 

In data analysis you will sort your data according to a certain variable in the data set. In R, this is done with the help of the function [`order()`](http://www.rdocumentation.org/packages/base/functions/order). 

[`order()`](http://www.rdocumentation.org/packages/base/functions/order) is a function that gives you the ranked position of each element when it is applied on a variable, such as a vector for example. Let us look at the vector `a`: `a <- c(100,9,101)`.

```
order(a)
``` 

returns the vector {2, 1, 3} since 100 is the second largest element of the vector, 9 is the smallest element and 101 is the largest element.
    
```
a[order(a)]
``` 
will thus give you the ordered vector {9, 100, 101}, since it first picks the second element of `a`, then the first and then the last. Got it? If you are not sure, use the console to play with the [`order()`](http://www.rdocumentation.org/packages/base/functions/order) function.` 

*** =instructions
Experiment with the [`order()`](http://www.rdocumentation.org/packages/base/functions/order) function in the console. Click 'Submit Answer' when you are ready to continue.

*** =hint
Just play with the [`order()`](http://www.rdocumentation.org/packages/base/functions/order) function in the console!

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Just play around with the order function in the console to see how it works!
```

*** =solution
```{r eval=FALSE}
# Just play around with the order function in the console to see how it works!
# Some examples:
order(1:10)
order(2:11)
order(c(5,4,6,7))
```

*** =sct
```{r eval=FALSE}
success_msg("Great! Now let's use the <code>order()</code> function to sort your data frame!")
```

---
## Sorting your data frame

Alright, now that you understand the [`order()`](http://www.rdocumentation.org/packages/base/functions/order) function, let us do something useful with it. You would like to rearrange your data frame such that it starts with the largest planet and ends with the smallest one. A sort on the `diameter` column, as you will.

*** =instructions
- Assign to the variable `positions` the desired ordering for the new data frame that you will create in the next step. You can use the [`order()`](http://www.rdocumentation.org/packages/base/functions/order) function for that, with the additional argument `decreasing = TRUE`.
- Now create the data frame `largest_first_df`, which contains the same information as `planets_df`, but with the planets in decreasing order of magnitude.

*** =hint
```
order(planets_df$diameter, decreasing = TRUE)
``` 
will give you the ordering of the variable diameter from largest to smallest. This is wat you should assign to positions. Use the variable positions then to select from the data frame `planets_df`!

*** =pre_exercise_code
```{r eval=FALSE}
planets <- c("Mercury", "Venus", "Earth", "Mars", "Jupiter", "Saturn", "Uranus", "Neptune");
type <- c("Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Terrestrial planet", "Gas giant", "Gas giant", "Gas giant", "Gas giant")
diameter <- c(0.382, 0.949, 1, 0.532, 11.209, 9.449, 4.007, 3.883); 
rotation <- c(58.64, -243.02, 1, 1.03, 0.41, 0.43, -0.72, 0.67);
rings <- c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE);
planets_df   <- data.frame(planets,type,diameter,rotation,rings)
```

*** =sample_code
```{r eval=FALSE}
# 'planets_df' is pre-loaded in your workspace

# What is the correct ordering based on the planets_df$diameter variable?
positions <-  

# Create new "ordered" data frame:
largest_first_df <-
```

*** =solution
```{r eval=FALSE}
# 'planets_df' is pre-loaded in your workspace

# What is the correct ordering based on the planets_df$diameter variable?
positions <- order(planets_df$diameter, decreasing = TRUE)

# Create new "ordered" data frame:
largest_first_df <- planets_df[positions, ]
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or overwrite the <code>planets_df</code> data frame!"
test_object("planets_df", undefined_msg = msg, incorrect_msg = msg)
test_object("positions",
            undefined_msg = "Please make sure to define a variable <code>positions</code>.",
            incorrect_msg = "It looks like <code>positions</code> does not contain all the correct ordering of the diameter column.")
test_object("largest_first_df",
            undefined_msg = "Please make sure to define a variable <code>largest_first_df</code>.",
            incorrect_msg = "It looks like <code>largest_first_df</code> does not contain the positions of the ordered <code>planets_df</code>.")
success_msg("Wonderful! What does the resulting data frame look like? Type <code>largest_first_df</code> in the console and find out! This exercise concludes the chapter on data frames. Remember that data frames are extremely important in R, you will need them all the time. Another very often used data structure is the list. This will be the subject of the next chapter!")
```