--- 
title_meta  : Chapter 6
title       : Lists
description : Lists, as opposed to vectors, can hold components of different types, just like your to-do list at home or at work. This intro to R chapter will teach you how to create, name and subset these lists.

--- type:NormalExercise xp:100 skills:1
## Lists, why would you need them?

Congratulations! At this point in the course you are already a (semi-)expert in:

- **Vectors** (one dimensional array): can hold numeric, character or logical values. The elements in one vector all have the same datatype.
- **Matrices** (two dimensional array): can hold numeric, character or logical values. The elements in one matrix all have the same datatype.
- **Data frames** (two-dimensional objects): can hold numeric, character or logical values. Within a column all elements have the same data type, but different columns can be of different data type.

Pretty sweet for an R newbie, right? ;-)

*** =instructions
Click 'Submit Answer' to start learning everything about lists!

*** =hint
Just click the 'Submit Answer' button.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Just click the 'Submit Answer' button.
```

*** =solution
```{r eval=FALSE}
# Just click the 'Submit Answer' button.
```

*** =sct
```{r eval=FALSE}
success_msg("Ready, set, go! Continue to the next exercise")
```


--- type:NormalExercise xp:100 skills:1
## Lists, why would you need them? (2)

A **list** in R is similar to your to-do list at work or school: the different items on that list most likely differ in length, characteristic, type of activity that has to do be done, ... 

A list in R allows you to gather a variety of objects under one name (that is, the name of the list) in an ordered way. These objects can be matrices, vectors, data frames, even other lists, etc. It is not even required that these objects are related to each other. Just like your to-do list :-). 

Maybe you can even say that a list is a kind of super data type ;-)

*** =instructions
Click 'Submit Answer' to start the first exercise on lists.

*** =hint
Click 'Submit Answer' to start the first exercise on lists.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Click 'Submit Answer' to start the first exercise on lists.
```

*** =solution
```{r eval=FALSE}
# Click 'Submit Answer' to start the first exercise on lists.
```

*** =sct
```{r eval=FALSE}
success_msg("Cool. Let's get our hands dirty!")
```


--- type:NormalExercise xp:100 skills:1
## Creating a list

Let us create our first list! To construct a list you use the function [`list()`](http://www.rdocumentation.org/packages/base/functions/list): 

```
my_list <- list(component1, component2 ...)
```

The arguments to the list function are the list components. Remember, these components can be a collection of matrices, vectors... 

*** =instructions
Construct a list, named `my_list`, that contains the variables `my_vector`, `my_matrix` and `my_df` as list components.

*** =hint
Use the [`list()`](http://www.rdocumentation.org/packages/base/functions/list) function with `my_vector`, `my_matrix` and `my_df` as arguments separated by a comma.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Vector with numerics from 1 up to 10
my_vector <- 1:10 
# Matrix with numerics from 1 up to 9
my_matrix <- matrix(1:9, ncol = 3)
# First 10 elements of the built-in data frame mtcars
my_df <- mtcars[1:10,]

# Construct list with these different elements:
my_list <- 
```

*** =solution
```{r eval=FALSE}
# Vector with numerics from 1 up to 10
my_vector <- 1:10 
# Matrix with numerics from 1 up to 9
my_matrix <- matrix(1:9, ncol = 3)
# First 10 elements of the built-in data frame mtcars
my_df <- mtcars[1:10,]

# Construct list with these different elements:
my_list <- list(my_vector, my_matrix, my_df)
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or change the definiton of the variables <code>my_vector</code>, <code>my_matrix</code> or <code>my_df</code>!"
test_object("my_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("my_matrix", undefined_msg = msg, incorrect_msg = msg)
test_object("my_df", undefined_msg = msg, incorrect_msg = msg)
test_object("my_list",
            undefined_msg = "Please make sure to define a variable <code>my_list</code>.",
            incorrect_msg = "It looks like <code>my_list</code> does not contain the correct elements.")
success_msg("Wonderful! Your skillset is growing at a staggering pace! Head over to the next exercise.")
```


--- type:NormalExercise xp:100 skills:1
## Creating a named list

Well done, you're on a roll!

Just like on your to-do list, you want to avoid not knowing or remembering what the components of your list stand for. That is why you should give names to them: 

```
my_list <- list(name1 = your_comp1, 
                name2 = your_comp2 ...)
``` 

This creates a list with the components `"name1"` and `"name2"` and so on. If you want to name your lists after you've created them, you can use the `names()` function as you did with vectors. The following commands are fully equivalent to the assignment above:

```
my_list <- list(your_comp1, your_comp2 ...)
names(my_list) <- c("name1", "name2" ...)
```

*** =instructions
- Change the code of the previous exercise (see editor) by adding names to the components. Use for `my_vector` the name `vec`, for `my_matrix` the name `mat` and for `my_df` the name `df`. 
- Print the list to the console and inspect the output.

*** =hint
The first method of assigning names to your list components is the easiest. It starts like this: 
```
my_list <- list(vec = my_vector)
```
Add the other two components in a similar fashion.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Vector with numerics from 1 up to 10
my_vector <- 1:10 
# Matrix with numerics from 1 up to 9
my_matrix <- matrix(1:9, ncol = 3)
# First 10 elements of the built-in data frame mtcars
my_df <- mtcars[1:10,]

# Construct my_list with these different elements:
my_list <- list(my_vector, my_matrix, my_df)

# Print my_list to the console

```

*** =solution
```{r eval=FALSE}
# Vector with numerics from 1 up to 10
my_vector <- 1:10 
# Matrix with numerics from 1 up to 9
my_matrix <- matrix(1:9, ncol = 3)
# First 10 elements of the built-in data frame mtcars
my_df <- mtcars[1:10,]

# Construct my_list with these different elements:
my_list <- list(vec = my_vector, mat = my_matrix, df = my_df)

# Print my_list to the console
my_list
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or change the definiton of the variables <code>my_vector</code>, <code>my_matrix</code> or <code>my_df</code>!"
test_object("my_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("my_matrix", undefined_msg = msg, incorrect_msg = msg)
test_object("my_df", undefined_msg = msg, incorrect_msg = msg)
test_object("my_list",
            undefined_msg = "Please make sure to define a variable <code>my_list</code>.",
            incorrect_msg = "It looks like <code>my_list</code> does not contain the correct elements.")
test_object("my_list", eq_condition = "equal",
            incorrect_msg = "It looks like <code>my_list</code> does not contain the correct naming for the components.");
test_output_contains("my_list", incorrect_msg = "Don't forget to print <code>my_list</code> to the console! Simply add <code>my_list</code> on a new line in the editor.")
success_msg("Great! Not only do you know how to construct lists now, you can also name them; a skill that will prove most useful in practice. Continue to the next exercise.")
```


--- type:NormalExercise xp:100 skills:1
## Creating a named list (2)

Being a huge movie fan (remember your job at LucasFilms), you decide to start storing information on good movies with the help of lists. 

Start by creating a list for the movie "The Shining". We have already created the variables `actors` and `reviews` in your R workspace. Type `actors` or `reviews` in the console to check these.

*** =instructions
Create the variable `shining_list`. The list contains the following components:
- moviename: "The Shining"
- actors: a vector with the main actors' names
- reviews: a data frame that contains some reviews

Do not forget to name the list components accordingly!

*** =hint
Let's get you started with a chunk of code:
```
shining_list <- list(moviename = "The Shining")
```
Can you complete the rest? You still have to add the actors (stored in the vector `actors`) and reviews (stored in the vector `reviews`)

*** =pre_exercise_code
```{r eval=FALSE}
actors      <- c("Jack Nicholson","Shelley Duvall","Danny Lloyd","Scatman Crothers","Barry Nelson")
sources     <- c("IMDb1","IMDb2","IMDb3")
comments    <- c("Best Horror Film I Have Ever Seen","A truly brilliant and scary film from Stanley Kubrick","A masterpiece of psychological horror")
scores      <- c(4.5,4,5)
reviews     <- data.frame(scores,sources,comments)
```

*** =sample_code
```{r eval=FALSE}
# The vectors actors and reviews are pre-loaded in the workspace

# Create the list shining_list
shining_list <- 
```

*** =solution
```{r eval=FALSE}
# The vectors actors and reviews are pre-loaded in the workspace

# Create the list shining_list
shining_list <- list(moviename = "The Shining", actors = actors, reviews = reviews)
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or change the definiton of the pre-set variables!"
test_object("reviews", undefined_msg = msg, incorrect_msg = msg)
test_object("actors", undefined_msg = msg, incorrect_msg = msg)
test_object("shining_list",
            undefined_msg = "Please make sure to define a variable <code>shining_list</code>.",
            incorrect_msg = "It looks like <code>shining_list</code> does not contain the correct elements.")
test_object("shining_list", eq_condition = "equal",
            incorrect_msg = "It looks like <code>shining_list</code> does not contain the correct naming for the components.");
success_msg("Wonderful! You now know how to construct and name lists. As in the previous chapters, let's look at how to select elements for lists. Head over to the next exercise")
```


--- type:NormalExercise xp:100 skills:1
## Selecting elements from a list

Your list will often be built out of numerous elements and components. Therefore, getting a single element, multiple elements, or a component out of it is not always straightforward. 

One way to select a component is using the numbered position of that component. For example, to "grab" the first component of `shining_list` you type 

```
shining_list[[1]]
```

A quick way to check this out is typing it in the console. Important to remember: to select elements out of a _data set_ you use square brackets `[ ]`, do not mix them up!

Another way to check is to refer to the names of the components. 

```
shining_list[["reviews"]]
``` 

selects the `reviews` data frame. The same is true for the shorter version `shining_list$reviews`, that uses the `$`-sign. 

Besides selecting components, you often need to select specific elements out of these components. For example, with `shining_list[[2]][1]` you select from the second component, `actors` (`shining_list[[2]]`), the first element (`[1]`). When you type this in the console, you will see the answer is Jack Nicholson.

*** =instructions
- Select from the `shining_list` the last actor and assign the result to `last_actor`.
- Select from the `shining_list` all information regarding the second review. Store the result in `second_review`.

*** =hint
- If you want to do things nicely: `length(shining_list$actors)` gives you the number of actors, and thus the element to select.
- You can select the information of the second review with `shining_list$reviews[2, ]`.

*** =pre_exercise_code
```{r eval=FALSE}
actors      <- c("Jack Nicholson","Shelley Duvall","Danny Lloyd","Scatman Crothers","Barry Nelson")
sources     <- c("IMDb1","IMDb2","IMDb3")
comments    <- c("Best Horror Film I Have Ever Seen","A truly brilliant and scary film from Stanley Kubrick","A masterpiece of psychological horror")
scores      <- c(4.5,4,5)
reviews     <- data.frame(scores,sources,comments)
shining_list <- list(moviename="The Shining",actors=actors, reviews=reviews)
```

*** =sample_code
```{r eval=FALSE}
# shining_list is already pre-loaded in the workspace

# Define last_actor
last_actor <- 

# Define second_review
second_review <- 
```

*** =solution
```{r eval=FALSE}
# shining_list is already pre-loaded in the workspace

# Define last_actor
last_actor <- shining_list$actors[length(shining_list$actors)]

# Define second_review
second_review <- shining_list$reviews[2,]
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or change the definiton of <code>shining_list</code>, which is pre-loaded in the workspace!"
test_object("shining_list", undefined_msg = msg, incorrect_msg = msg)
test_object("last_actor", 
            undefined_msg = "Please make sure to define a variable <code>last_actor</code>.",
            incorrect_msg = "Looks like <code>last_actor</code> does not equal the last actor.")
test_object("second_review",
            undefined_msg = "Please make sure to define a variable <code>second_review</code>.",
            incorrect_msg = "It looks like <code>second_review</code> does not contain all the information regarding the second review.")
success_msg("Great! Selecting elements from lists is rather easy isn't it? Continue to the next exercise.")
```


--- type:NormalExercise xp:100 skills:1
## Adding more movie information to the list

Being proud of your first list, you shared it with the members of your movie hobby club. However, one of the senior members, a guy named M. McDowell, noted that you forgot to add the release year. Given your ambitions to become next year's president of the club, you decide to add this information to the list. 

In order to conveniently add elements to lists you use the concatenate function [`c()`](http://www.rdocumentation.org/packages/base/functions/c): 

```
c(list1 , some_variable)
``` 

This will simply extend the original list, `list1`, with the component `some_variable`. This component gets appended to the end of the list.
If you want to give the new list item a name, you just add the name as you did before: 

```
c(list1, new_name = some_variable)
```

*** =instructions
- Complete the code below such that an item named `year` is added to the `shining_list` with the value 1980. Assign the result to `shining_list_full`.
- Finally, have a look at the structure of `shining_list_full` with the [`str()`](http://www.rdocumentation.org/packages/utils/functions/str) function.

*** =hint
Have a look at the example code in the exercise assignment. Maybe this can help you start:
```
shining_list <- c(shining_list, ...)
```
You still have to add some code where the three dots are.

*** =pre_exercise_code
```{r, eval = FALSE}
actors      <- c("Jack Nicholson","Shelley Duvall","Danny Lloyd","Scatman Crothers","Barry Nelson")
sources     <- c("IMDb1","IMDb2","IMDb3")
comments    <- c("Best Horror Film I Have Ever Seen","A truly brilliant and scary film from Stanley Kubrick","A masterpiece of psychological horror")
scores      <- c(4.5,4,5)
reviews     <- data.frame(scores,sources,comments)
shining_list <- list(moviename="The Shining",actors=actors, reviews=reviews)
```

*** =sample_code

```{r eval=FALSE}
# shining_list, the list containing moviename, actors and reviews, is pre-loaded in the workspace

# We forgot something; add the year to shining_list
shining_list_full <- 

# Have a look at shining_list_full

```

*** =solution

```{r eval=FALSE}
# shining_list, the list containing moviename, actors and reviews, is pre-loaded in the workspace

# We forgot something; add the year to shining_list:
shining_list_full <- c(shining_list, year = 1980)

# Have a look at shining_list_full
str(shining_list_full)
```

*** =sct
```{r eval=FALSE}
msg = "Do not remove or change the definiton of <code>shining_list</code>, which is pre-loaded in the workspace!"
test_object("shining_list", undefined_msg = msg, incorrect_msg = msg)
test_object("shining_list_full",
            undefined_msg = "Please make sure to define a variable <code>shining_list_full</code>.",
            incorrect_msg = "It looks like <code>shining_list_full</code> does not contain all the correct elements. It should contain the <code>year</code> (1980) as well!")
test_object("shining_list_full", eq_condition = "equal",
            incorrect_msg = "It looks like <code>shining_list_full</code> does not contain the correct naming for all the components.");
test_function("str", "object", incorrect_msg = "Make sure to inspect the structure of <code>shining_list_full</code>.")
success_msg("Great! This was the last exercise on R lists! You now have a solid basis in the R programming language, but there's so much more to learn. Check out all the other courses and DataCamp and become a true data science expert!")
```


