--- 
title_meta  : Chapter 2
title       : Vectors
description : We take you on a trip to Vegas, where you will learn how to analyze your gambling results using vectors in R! After completing this chapter, you will be able to create vectors in R, name them, select elements from them and compare different vectors. 
framework   : datacamp
mode        : selfcontained
---

## Create a vector

Feeling lucky? 

You better, because this chapter takes you on a trip to the City of Sins, also known as *"Statisticians Paradise"* ;-).

Thanks to R and your new data-analytical skills, you will learn how to uplift your performance at the tables and fire off your career as a professional gambler. This chapter will show how you can easily keep track of your betting progress and how you can do some simple analyses on past actions. Next Stop, Vegas Baby... VEGAS!!

*** =instructions
- Do you still remember what you have learned in the first chapter? Assign the value `"Here we go!"` to the variable `Vegas`!

*** =hint
Just type the following line in the editor:
```
Vegas <- "Here we go!"
```
Note that R is case sensitive.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Define the variable 'Vegas'
```

*** =solution
```{r eval=FALSE}
# Define the variable 'Vegas'
Vegas <- "Here we go!"
```

*** =sct
```{r eval=FALSE}
test_object("Vegas", 
            undefined_msg = "Please make sure to define a variable <code>Vegas</code>.",
            incorrect_msg = "Make sure that you assign the correct value to <code>Vegas</code>. Do not forget that R is case sensitive!")
success_msg("Great! Head over to the next exercise.")
```

---

## Create a vector (2)

Let us focus first! 

On your way from rags to riches, you will make extensive use of vectors. Vectors are one-dimension arrays that can hold numeric data, character data, or logical data. In other words, a vector is a simple tool to store data. For example, you can store your daily gains and losses in the casinos. 

In R, you create a vector with the combine function [`c()`](http://www.rdocumentation.org/packages/base/functions/c). You place the vector elements separated by a comma between the brackets. For example:

```
numeric_vector <- c(1, 2, 3)
character_vector <- c("a", "b", "c")
boolean_vector <- c(TRUE, FALSE)
```

Once you have created these vectors in R, you can use them to do calculations.

*** =instructions 
Complete the code such that `boolean_vector` contains the three elements: `TRUE`, `FALSE` and `TRUE` (in that order). 

*** =hint 
Assign `c(TRUE,FALSE,TRUE)` to the variable `boolean_vector` with the `<-` operator. You can also use the shorter notations, `T` and `F` for `TRUE` and `FALSE` respectively.


*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")
# Complete the code for 'boolean_vector'
boolean_vector <-
```

*** =solution
```{r eval=FALSE}
numeric_vector <- c(1, 10, 49)
character_vector <- c("a", "b", "c")
# Complete the code for 'boolean_vector'
boolean_vector <- c(TRUE, FALSE, TRUE)
```

*** =sct
```{r eval=FALSE}
test_object("numeric_vector")
test_object("character_vector")
test_object("boolean_vector",
            undefined_msg = "Please make sure to define a variable <code>boolean_vector</code>.",
            incorrect_msg = "Make sure that you assign the correct values to <code>boolean_vector</code>. The order matters!")
success_msg("Perfect! Notice that adding a space behind the commas in the <code>c()</code> function improve the readability of your code. Let's practice some more with vector creation in the next exercise.")
```

---

## Create a vector (3)

After one week in Las Vegas and still zero Ferraris in your garage, you decide that it is time to start using your data analytical superpowers.

Before doing a first analysis, you decide to first collect all the winnings and losses for the last week: 

For `poker_vector`: 
- On Monday you won 140\$ 
- Tuesday you lost 50\$
- Wednesday you won 20\$ 
- Thursday you lost 120\$ 
- Friday you won 240\$

For `roulette_vector`: 
- On Monday you lost 24\$ 
- Tuesday you lost 50\$
- Wednesday you won 100\$ 
- Thursday you lost 350\$
- Friday you won 10\$

You only played poker and roulette, since there was a delegation of mediums that occupied the craps tables. To be able to use this data in R, you decide to create the variables `poker_vector` and `roulette_vector`.

*** =instructions
Assign the winnings/losses for roulette to the variable `roulette_vector`.

*** =hint
To help you with this step, the editor already contains the code for creating `poker_vector`. Assign the correct values to `roulette_vector` based on the numbers in the assignment. Do not forget that losses are negative numbers.

*** =sample_code

```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <-  
```

*** =solution

```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)
```

*** =sct
```{r eval=FALSE}
test_object("poker_vector",
            undefined_msg = "Please make sure to define a variable <code>poker_vector</code>.",
            incorrect_msg = "Make sure that you assign the correct values to <code>poker_vector</code>.")
test_object("roulette_vector",
            undefined_msg = "Please make sure to define a variable <code>roulette_vector</code>.",
            incorrect_msg = "Make sure that you assign the correct values to <code>roulette_vector</code>.")
success_msg("Very good! To check out the contents of your vectors, remember that you can always simply type the variable in the console and hit Enter. Proceed to the next exercise!")
```

--- 

## Naming a vector

As a data analyst, it is important to have a clear view on the data that you are using. Understanding what each element refers to is therefore essential. 

In the previous exercise, we created a vector with your winnings over the week. Each vector element refers to a day of the week but it is hard to tell which element belongs to which day. It would be nice if you could show that in the vector itself. 

You can give a name to the elements of a vector with the `names()` function. Have a look at this example:

```{r}
some_vector <- c("Johnny", "Poker Player")
names(some_vector) <- c("Name", "Profession")
```

This code first creates a vector `some_vector` and then gives the two elements a name. The first element is assigned the name `Name`, while the second element is labeled `Profession`. Printing the contents to the console yields following output:

```{r, echo = FALSE}
some_vector
```

*** =instructions
Go ahead and assign the days of the week as names to `poker_vector` and `roulette_vector`. In case you are not sure, the days of the week are: Monday, Tuesday, Wednesday, Thursday and Friday.

*** =hint
You can use `names(poker_vector)` to set the names of the variable `poker_vector`. Make sure not to forget the parentheses and the fact that weekdays are written with a capital letter! The vector with days of the week is thus given by 
```
c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
```

*** =pre_exercise_code
```{r eval=FALSE}
```

*** =sample_code
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Add your code here

```

*** =solution
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Add your code here
names(poker_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
```

*** =sct
```{r eval=FALSE}
test_object("poker_vector",
            undefined_msg = "Please make sure to define a variable <code>poker_vector</code>.",
            incorrect_msg = "Do not change the values inside <code>poker_vector</code>; they were already coded for you.")
test_object("roulette_vector",
            undefined_msg = "Please make sure to define a variable <code>roulette_vector</code>.",
            incorrect_msg = "Do not change the values inside <code>roulette_vector</code>; they were already coded for you.")
test_object("poker_vector",
            eq_condition = "equal",
            incorrect_msg = "Make sure that you assign the correct names vector to <code>poker_vector</code>.")
test_object("roulette_vector",
            eq_condition = "equal",
            incorrect_msg = "Make sure that you assign the correct names vector to <code>roulette_vector</code>.")
success_msg("Well done! Continue to the next exercise.")
```

---
## Naming a vector (2)

If you want to become a good statistician, you have to become lazy. (If you are already lazy, chances are high you are one of those exceptional, natural-born statistical talents.)

In the previous exercises you probably experienced that it is boring and frustrating to type and retype information such as the days of the week. However, when you look at it from a higher perspective, there is a more efficient way to do this, namely, to assign the days of the week vector to a **variable**! 

Just like you did with your poker and roulette returns, you can also create a variable that contains the days of the week. This way you can use and re-use it.

*** =instructions
- Create a variable `days_vector` that contains the days of the week, from Monday to Friday.
- Use that variable `days_vector` to set the names of `poker_vector` and `roulette_vector`.

*** =hint
You can use `names(poker_vector)` to set the names of the variable `poker_vector`. Make sure not to forget the parentheses around the days of the week.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Create the variable 'days_vector'
days_vector <-
 
#Assign the names of the day to 'roulette_vector' and 'poker_vector'
names(poker_vector) <-   
names(roulette_vector) <-
```

*** =solution
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Create the variable 'days_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

#Assign the names of the day to 'roulette_vector' and 'poker_vector'
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector
```

*** =sct
```{r eval=FALSE}
test_object("poker_vector",
            undefined_msg = "Please make sure to define a variable <code>poker_vector</code>.",
            incorrect_msg = "Do not change the values inside <code>poker_vector</code>; they were already coded for you.")
test_object("roulette_vector",
            undefined_msg = "Please make sure to define a variable <code>roulette_vector</code>.",
            incorrect_msg = "Do not change the values inside <code>roulette_vector</code>; they were already coded for you.")
test_object("days_vector",
            undefined_msg = "Please make sure to define a variable <code>days_vector</code>.",
            incorrect_msg = "Make sure that you assign the correct vector to <code>days_vector</code>.")
test_object("poker_vector",
            eq_condition = "equal",
            incorrect_msg = "Make sure that you assign <code>days_vector</code> to the names of <code>poker_vector</code>.")
test_object("roulette_vector",
            eq_condition = "equal",
            incorrect_msg = "Make sure that you assign <code>days_vector</code> to the names of <code>roulette_vector</code>.")
success_msg("Nice one! A word of advice: try to avoid code duplication at all times. Continue to the next exercise and learn how to do arithmetic with vectors!")
```

---

## Calculating total winnings

Now that you have the poker and roulette winnings nicely as a named vector, you can start doing some data analytical magic. 

You want to find out the following type of information:
- How much has been your overall profit or loss per day of the week?
- Have you lost money over the week in total?
- Are you winning/losing money on poker or on roulette?

To get the answers, you have to do arithmetic calculations on vectors. 

It is important to know is that if you sum two vectors in R, it takes the element-wise sum. For example, the following three statements are completely equivalent:

```
c(1, 2, 3) + c(4, 5, 6)
c(1 + 4, 2 + 5, 3 + 6)
c(5, 7, 9)
```

Let us try this first!

*** =instructions
- Take the sum of the variables `A_vector` and `B_vector` and it assign to `total_vector`.
- Inspect the result by printing `total_vector` to the console.

*** =hint
Use the `+` operator to sum `A_vector` and `B_vector`. Use `=` to assign the result to `total_vector`.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
A_vector <- c(1, 2, 3)
B_vector <- c(4, 5, 6)

# Take the sum of 'A_vector' and 'B_vector'
total_vector <- 
  
# Print 'total_vector' to the console
total_vector
```

*** =solution
```{r eval=FALSE}
A_vector <- c(1, 2, 3)
B_vector <- c(4, 5, 6)

# Take the sum of A_vector and B_vector
total_vector <- A_vector + B_vector

# Print 'total_vector' to the console
total_vector
```

*** =sct
```{r eval=FALSE}
test_object("A_vector",
            undefined_msg = "Please make sure to define a variable <code>A_vector</code>.",
            incorrect_msg = "Make sure that you assign the correct vector to <code>A_vector</code>.")
test_object("B_vector",
            undefined_msg = "Please make sure to define a variable <code>B_vector</code>.",
            incorrect_msg = "Make sure that you assign the correct vector to <code>B_vector</code>.")
test_object("total_vector",
            undefined_msg = "Please make sure to define a variable <code>total_vector</code>.",
            incorrect_msg = "Make sure that <code>total_vector</code> contains the sum of <code>A_vector</code> and <code>B_vector</code>.")
success_msg("Good job! Continue to the next exercise.")
```

---

## Calculating total winnings (2)

Do you understand how R does arithmetic calculations with vectors? 

Then it is time to get those Ferraris in your garage! First, you need to understand what the overall profit or loss per day of the week was. The total daily profit is the sum of the profit/loss you realized on poker per day, and the profit/loss you realized on roulette per day. 

In R, this is just the sum of `roulette_vector` and `poker_vector`.

*** =instructions
Assign to the variable `total_daily` how much you won or lost on each day in total (poker and roulette combined).

*** =hint
Similar to the previous exercise, assign the sum of two vectors to a new variable, `total_daily`.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Poker winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday:
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Up to you now:
total_daily <- 
```

*** =solution
```{r eval=FALSE}
# Poker winnings from Monday to Friday:
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday:
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(poker_vector) <- days_vector
names(roulette_vector) <- days_vector

# Up to you now:
total_daily <- poker_vector + roulette_vector
```

*** =sct
```{r eval=FALSE}
msg = "Do not change anything about the definition and naming of <code>poker_vector</code> and <code>roulette_vector</code>."
test_object("days_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("total_daily", 
            incorrect_msg = "Make sure that you assign the sum of <code>poker_vector</code> and <code>roulette_vector</code> to <code>total_daily</code>.")

success_msg("Great! Continue to the next exercise.")
```

---

## Calculating total winnings (3)

Based on the previous analysis, it looks like you had a mix of good and bad days. This is not what your ego expected, and you wonder if there may be a (very very very) tiny chance you have lost money over the week in total? 

A function that helps you to answer this question is [`sum()`](http://www.rdocumentation.org/packages/base/functions/sum). It calculates the sum of all elements of a vector. For example, to calculate the total amount of money you have lost/won with poker you do: 

```
total_poker <- sum(poker_vector)
```

*** =instructions
- Calculate the total amount of money that you have won/lost with roulette and assign to the variable `total_roulette`.
- Now that you have the totals for roulette and poker, you can easily calculate `total_week` (which is the sum of all gains and losses of the week).

*** =hint
Use the `sum` function to get the total of the `roulette_vector`. `total_week` is then the sum of `roulette_vector` and `poker_vector`.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# Total winnings with poker
total_poker <- sum(poker_vector)

# Up to you now:
total_roulette <-
total_week <- 
```

*** =solution
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

total_poker <- sum(poker_vector)

# Up to you now:
total_roulette <-  sum(roulette_vector)
total_week <- total_roulette + total_poker
```

*** =sct
```{r eval=FALSE}
msg = "Do not change anything about the definition and naming of <code>poker_vector</code> and <code>roulette_vector</code>."
test_object("days_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("total_poker", 
            undefined_msg = "Please make sure to define a variable <code>total_poker</code>.",
            incorrect_msg = "Make sure that you assign to <code>total_poker</code> the sum of the <code>poker_vector</code>.")
test_object("total_roulette",
            undefined_msg = "Please make sure to define a variable <code>total_roulette</code>.",
            incorrect_msg = "Make sure that you assign to <code>total_roulette</code> the sum of the <code>roulette_vector</code>.")
test_object("total_week",
            undefined_msg = "Please make sure to define a variable <code>total_week</code>.",
            incorrect_msg = "Make sure that you assign to <code>total_week</code> the sum of the other two total vectors.")
```

---

## Comparing total winnings

Oops, it seems like you are losing money. Time to rethink and adapt your strategy! This will require some deeper analysis... 

After a short brainstorm in your hotel's jacuzzi, you realize that a possible explanation might be that your skills in roulette are not as well developed as your skills in poker. So maybe your total gains in poker are higher (or `>` ) than in roulette.

*** =instructions
- Calculate `total_poker` and `total_roulette` as in the previous exercise.
- Check if your total gains in poker are higher than for roulette by using a comparison. Assign the result of this comparison to the variable `answer`.
What do you conclude, should you focus on roulette or on poker?

*** =hint
You partly calculated the answer to this question in the previous exercise already!
To check if 6 is larger than 5, you type `6 > 5`. This returns a variable with data type Logical (`TRUE` or `FALSE`).

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# Calculate total gains for poker and roulette
total_poker <-
total_roulette <-

# Check if you realized higher total gains in poker than in roulette 
answer <-
```

*** =solution
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# Calculate total gains for poker and roulette
total_poker <- sum(poker_vector)
total_roulette <- sum(roulette_vector)

# Check if you realized higher total gains in poker than in roulette
answer <- total_poker > total_roulette
```

*** =sct
```{r eval=FALSE}
msg = "Do not change anything about the definition and naming of <code>poker_vector</code> and <code>roulette_vector</code>."
test_object("days_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("total_poker", 
            undefined_msg = "Please make sure to define a variable <code>total_poker</code>.",
            incorrect_msg = "Make sure that you assign to <code>total_poker</code> the sum of the <code>poker_vector</code>.")
test_object("total_roulette",
            undefined_msg = "Please make sure to define a variable <code>total_roulette</code>.",
            incorrect_msg = "Make sure that you assign to <code>total_roulette</code> the sum of the <code>roulette_vector</code>.")
test_object("answer",
            undefined_msg = "Please make sure to define a variable <code>answer</code>.",
            incorrect_msg = "It looks like the answer variable does not contain the correct answer. Make sure to read the assignment carefully.")
success_msg("Good job! Continue to the next exercise.")
```

---

## Vector selection: the good times

Your hunch seemed to be right. It appears that the poker game is more your cup of tea than roulette. 

Another possible route for investigation is your performance at the beginning of the working week compared to the end of it. You did have a couple of Margarita cocktails at the end of the week... 

To answer that question, you only want to focus on a selection of the `total_vector`. In other words, our goal is to select specific elements of the vector. To select elements of a vector (and later matrices, data frames, ...), you can use square brackets. Between the square brackets, you indicate what elements to select. For example, to select the first element of the vector, you type `poker_vector[1]`. To select the second element of the vector, you type `poker_vector[2]`, etc.

*** =instructions
Assign the poker results of Wednesday to the variable `poker_wednesday`.

*** =hint
Wednesday is the third element of `poker_vector`, and can thus be selected with `poker_vector[3]`.
Notice that the first element in a vector has index 1, not 0 as in many other programming languages.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# Define a new variable based on a selection
poker_wednesday <- 
```

*** =solution
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# Define a new variable based on a selection
poker_wednesday <- poker_vector[3]
```

*** =sct
```{r eval=FALSE}
msg = "Do not change anything about the definition and naming of <code>poker_vector</code> and <code>roulette_vector</code>."
test_object("days_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_wednesday", 
            undefined_msg = "Please make sure to define a variable <code>poker_wednesday</code>.",
            incorrect_msg = "It looks like <code>poker_wednesday</code> does not contain the correct value of the <code>poker_vector</code>.")
success_msg("Great! R also makes it possible to select multiple elements from a vector at once. Learn how in the next exercise!")
```

---
## Vector selection: the good times (2)

How about analyzing your midweek results? 

To select multiple elements from a vector, you can add square brackets at the end of it. You can indicate between the brackets what elements should be selected. For example: suppose you want to select the first and the fifth day of the week: use the vector `c(1,5)` between the square brackets. For example, the code below selects the first and fifth element of `poker_vector`:

```
poker_vector[c(1,5)]
```

*** =instructions
Assign the poker results of Tuesday, Wednesday and Thursday to the variable `poker_midweek`.

*** =hint
Use the vector `c(2,3,4)` between square brackets to select the correct elements of `poker_vector`.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
``` 

*** =sample_code
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# Define a new variable based on a selection
poker_midweek <- 
```

*** =solution
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# Define a new variable based on a selection
poker_midweek <- poker_vector[c(2, 3, 4)]
```

*** =sct
```{r eval=FALSE}
msg = "Do not change anything about the definition and naming of <code>poker_vector</code> and <code>roulette_vector</code>."
test_object("days_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_midweek", 
            undefined_msg = "Please make sure to define a variable <code>poker_midweek</code>.",
            incorrect_msg = "It looks like <code>poker_midweek</code> does not contain the correct values from <code>poker_vector</code>.")
success_msg("Well done! Continue to the next exercise to specialize in vector selection some more!");
```

---

## Vector selection: the good times (3)

Selecting multiple elements of `poker_vector` with `c(2,3,4)` is not very convenient. Many statisticians are lazy people by nature, so they created an easier way to do this: `c(2,3,4)` can be abbreviated to`2:4`, which generates a vector with all natural numbers from 2 up to 4.

So, another way to find the mid-week results is `poker_vector[2:4]`. Notice how the vector `2:4` is placed between the square brackets to select element 2 up to 4.

*** =instructions
Assign the results to `roulette_selection_vector` the results from Tuesday up to Friday by making use of `:`.

*** =hint
Assign a selection of `roulette_vector` to `roulette_selection_vector` by placing `2:5` between square brackets.

*** =sample_code

```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# Define a new variable based on a selection
roulette_selection_vector <- 
```

*** =solution

```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# Define a new variable based on a selection
roulette_selection_vector <- roulette_vector[2:5]
```

*** =sct

```{r eval=FALSE}
msg = "Do not change anything about the definition and naming of <code>poker_vector</code> and <code>roulette_vector</code>."
test_object("days_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_selection_vector", 
            undefined_msg = "Please make sure to define a variable <code>roulette_selection_vector</code>.",
            incorrect_msg = "It looks like <code>roulette_selection_vector</code> does not contain the correct selection from <code>roulette_vector</code>. Make sure to to use the right indices.")
success_msg("Awesome! The colon operator is extremely useful and very often used in R programming, so remember it well. Proceed to the next exercise.")
```

---

## Vector selection: the good times (4)

Another way to tackle the previous exercise is by using the names of the vector elements (Monday, Tuesday, ...) instead of their numeric positions. For example, 

```
poker_vector["Monday"]
```

will select the first element of `poker_vector` since `"Monday"` is the name of that first element.

Just like you did in the previous exercise with numerics, you can also use the element names to select multiple elements, for example: 

```
poker_vector[c("Monday","Tuesday")]
```

*** =instructions
Calculate your average poker gains during the first three days of the week by selecting these elements with the help of the names. Assign this value to `average_midweek_gain`. You can use the [`mean()`](http://www.rdocumentation.org/packages/base/functions/mean) function to get the average of a vector.

*** =hint
You can use `mean(poker_vector)` to get the mean of the vector poker. You do not need the mean of all poker elements, but only of the first three days.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

average_midweek_gain <- 
```

*** =solution
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

average_midweek_gain <- mean(poker_vector[c("Monday", "Tuesday", "Wednesday")])
```

*** =sct
```{r eval=FALSE}
msg = "Do not change anything about the definition and naming of <code>poker_vector</code> and <code>roulette_vector</code>."
test_object("days_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("average_midweek_gain", 
            undefined_msg = "Please make sure to define a variable <code>average_midweek_gain</code>.",
            incorrect_msg = "It looks like <code>average_midweek_gain</code> does not contain the mean of the correct selection of <code>poker_vector</code>.")
success_msg("Good job! Apart from subsetting vectors by index or by name, you can also subset vectors by comparison. The next exercises will show you how!");
```

---

## Selection by comparison - Step 1

By making use of comparison operators, we can approach the previous question in a more pro-active way. 

The (logical) comparison operators known to R are:
- `<` for less than
- `>` for greater than
- `>=` for greater than or equal to
- `==` for equal to each other
- `!=` not equal to each other

As seen in the previous chapter, stating `6 > 5` returns `TRUE`. The nice thing about R is that you can use these comparison operators also on vectors. For example, the statement `c(4,5,6) > 5` returns: `FALSE FALSE TRUE`. In other words, you test for every element of the vector if the condition stated by the comparison operator is `TRUE` or `FALSE`. Do not take our word for it! Try it in the console ;-).

Behind the scenes, R *recycles* the value `5` when you execute `c(4,5,6) > 5`. R wants to do an element-wise comparison of each element in `c(4,5,6)` with each corresponding element in `5`. However, `5` is not a vector of length three. To solve this, R automatically replicates the value 5 to generate a vector of three elements, `c(5, 5, 5)` and then carries out the element-wise comparison.

*** =instructions
Check if your poker winnings are positive on the different days of the week (i.e. > 0), and assign this to `selection_vector`.

*** =hint
In order to check for which days your poker gains are positive, R should check for each element of `poker_vector` whether it is larger than zero. `some_vector > 0` is the way to tell R what you are after. R will recycle the value `0` to generate a vector of zeros that has the same length as `some_vector`.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# What days of the week did you make money on poker?
selection_vector <- 
```

*** =solution
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# What days of the week did you make money on poker?
selection_vector <- poker_vector > 0
```

*** =sct
```{r eval=FALSE}
msg = "Do not change anything about the definition and naming of <code>poker_vector</code> and <code>roulette_vector</code>."
test_object("days_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("selection_vector", 
            undefined_msg = "Please make sure to define a variable <code>selection_vector</code>.",
            incorrect_msg = "It looks like <code>selection_vector</code> does not contain the correct result. Remember that R uses element wise operations for vectors.")
success_msg("Great! Instead of having R recycle the value <code>0</code> five times and then do the comparison, you can of course also compare <code>poker_vector</code> with a vector of length 5 that you've defined yourself. In that case, R will simply carry out the element-wise comparison without recycling anything.")
```

---

## Selection by comparison - Step 2

Working with comparisons will make your data analytical life easier. Instead of selecting a subset of days to investigate yourself (like before), you can simply ask R to return only those days where you realized a positive return for poker. 

In the previous exercises you used `selection_vector <- poker_vector > 0` to find the days on which you had a positive poker return. Now, you would like to know not only the days on which you won, but also how much you won on those days. 

You can select the desired elements, by putting `selection_vector` between square brackets when selecting from `poker_vector`. This works, because R only selects those elements where `selection_vector` is `TRUE` by default. For `selection_vector` this means where `poker_vector > 0`.

*** =instructions
Assign the amounts that you won on the profitable days to the variable `poker_winning_days`.

*** =hint
Use `selection_vector` to select the desired elements from `poker_vector`, and assign the result to `poker_winning_days`.

*** =pre_exercise_code
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# What days of the week did you make money on poker?
selection_vector <- poker_vector > 0

# Select from poker_vector these days
poker_winning_days <- 
```

*** =solution
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# What days of the week did you make money on poker?
selection_vector <- poker_vector > 0

# Select from poker_vector these days
poker_winning_days <- poker_vector[selection_vector]
```

*** =sct
```{r eval=FALSE}
msg = "Do not change anything about the definition and naming of <code>poker_vector</code> and <code>roulette_vector</code>."
test_object("days_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("selection_vector", 
            undefined_msg = "Please make sure to define a variable <code>selection_vector</code>.",
            incorrect_msg = "It looks like <code>selection_vector</code> does not contain the correct result. Remember that R uses element wise operations for vectors.")
test_object("poker_winning_days",
            undefined_msg =  "Please make sure to define a variable <code>poker_winning_days</code>.",
            incorrect_msg =  "It looks like <code>poker_winning_days</code> does not contain the correct result. Remember that R uses element wise operations for vectors.")
success_msg("Good job! Continue to the next exercise.")
```

---

## Advanced selection

Just like you did for poker, you also want to know those days where you realized a positive return for roulette.

*** =instructions
- Assign the amounts that you made on the days that you ended positively for roulette to the variable `roulette_winning_days`. This vector thus contains the positive winnings of `roulette_vector`.

*** =hint
Have a look at the previous exercise and do the similar analysis, this time for `roulette_vector`.


*** =pre_exercise_ocde
```{r eval=FALSE}
# no pec
```

*** =sample_code
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# What days of the week did you make money on roulette?
selection_vector <-

# Select from roulette_vector these days

```

*** =solution
```{r eval=FALSE}
# Poker winnings from Monday to Friday
poker_vector <- c(140, -50, 20, -120, 240)

# Roulette winnings from Monday to Friday
roulette_vector <- c(-24, -50, 100, -350, 10)

# Give names to both 'poker_vector' and 'roulette_vector'
days_vector <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
names(roulette_vector) <- days_vector
names(poker_vector) <- days_vector

# What days of the week did you make money on roulette?
selection_vector <- roulette_vector > 0

# Select from roulette_vector these days
roulette_winning_days <- roulette_vector[selection_vector]
```

*** =sct
```{r eval=FALSE}
msg = "Do not change anything about the definition and naming of <code>poker_vector</code> and <code>roulette_vector</code>."
test_object("days_vector", undefined_msg = msg, incorrect_msg = msg)
test_object("poker_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("roulette_vector", eq_condition = "equal", undefined_msg = msg, incorrect_msg = msg)
test_object("selection_vector", 
            undefined_msg = "Please make sure to define a variable <code>selection_vector</code>.",
            incorrect_msg = "It looks like <code>selection_vector</code> does not contain the correct result. Remember that R uses element wise operations for vectors.")
test_object("roulette_winning_days",
            undefined_msg = "Please make sure to define a variable <code>roulette_winning_days</code>.",
            incorrect_msg = "It looks like <code>roulette_winning_days</code> does not contain the correct result. Make sure that you use <code>roulette_vector</code>.")
success_msg("Great! This exercise concludes the chapter on vectors. The next chapter will introduce you to the two-dimensional version of vectors: matrices.")
```
